---
title: "0.2.- Identify relatives on the UK Biobank"
---

```{r setup}
source('0.0.- Setup.R')
```

```{r read_demographics}
ukb_demo <- read_csv(file.path(root_fd, 'data', 'pheno', 'ukb_demo.csv'),
                     col_types = cols())
```

```{r read_cn}
ukb_47s <- read_csv(file.path(root_fd, 'data', 'cn', 'ukb_ratios.csv'),
                    col_types = cols()) |>
  filter(rdna > 0) |>
  rename(CN = rdna_ratio)
```

```{r find_cn_wb}
ukb_47s_wb <- ukb_47s |>
  inner_join(ukb_demo |>
               select(eid, Eth_Back_Gen, Eth_Back),
             by = 'eid') |>
  filter(Eth_Back_Gen == 1, Eth_Back == 1001) |>
  select(-starts_with('Eth_Back'))
```

```{r read_relatedness}
ukb_rel <- read_delim(
  file.path(root_fd, 'data', 'metadata', 'ukb_rel.txt'),
  col_types = cols(), delim=' ') |>
  mutate(relation=row_number(),
         dist = case_when(
           Kinship > 0.4 ~ 'MZ Twins',
           Kinship < 0.354 & Kinship >= 0.177  ~ 'First-Degree',
           Kinship < 0.177 & Kinship >= 0.0884 ~ 'Second-Degree',
           Kinship < 0.0884 & Kinship >= 0.0442 ~ 'Third-Degree',
           TRUE ~ 'Unknown'))
```

```{r save_relatives_distance}
ukb_rel |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_rel_dist.csv'),
    na = '')
```

## All relatives

```{r assign_age_relatives}
ukb_rel_ages <-  ukb_rel |>
  select(eid_1 = ID1, eid_2 = ID2, relation, dist) |>
  inner_join(ukb_demo |> select(eid, Age, Birth_Year),
             by = c('eid_1' = 'eid')) |>
  inner_join(ukb_demo |> select(eid, Age, Birth_Year),
             by = c('eid_2' = 'eid'),
             suffix = c('_1', '_2')) |>
  mutate(
    eid_o = if_else(Birth_Year_2 > Birth_Year_1, eid_1, eid_2),
    eid_y = if_else(eid_o == eid_1, eid_2, eid_1),
    Age_o = if_else(Birth_Year_2 > Birth_Year_1, Age_1, Age_2),
    Age_y = if_else(eid_o == eid_1, Age_2, Age_1),
    Birth_Year_o = if_else(Birth_Year_2 > Birth_Year_1, 
                           Birth_Year_1, Birth_Year_2),
    Birth_Year_y = if_else(eid_o == eid_1, Birth_Year_2, Birth_Year_1)) |>
  select(relation, dist, ends_with(c('_o', 'y')))
```

```{r save_relatives_age}
ukb_rel_ages |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_rel_dist_age.csv'),
    na = '')
```

## First degree relatives

IMPORTANT: 
Removing "parental" pairs < 10 years of difference

```{r identify_first_degree_relatives}
ukb_first_pheno <- ukb_rel |>
  filter(grepl('First', dist)) |>
  inner_join(
    ukb_demo |>
      select(eid,
             Sex_1=Sex,
             Birth_Year_1=Birth_Year),
    by=c('ID1'='eid')) |>
  inner_join(
    ukb_demo |>
      select(eid,
             Sex_2=Sex,
             Birth_Year_2=Birth_Year),
    by=c('ID2'='eid'))
```

```{r classify_first_degree_relatives}
ukb_first_types <- ukb_first_pheno |>
  mutate(the_diff = Birth_Year_2 - Birth_Year_1,
         rel_type = if_else(
           IBS0 < 0.0012, 'Parental', 'Fraternal'),
         eid_o = if_else(the_diff > 0, ID1, ID2),
         eid_y = if_else(the_diff > 0, ID2, ID1),
         Sex_o = if_else(the_diff > 0, Sex_1, Sex_2),
         Sex_y = if_else(the_diff > 0, Sex_2, Sex_1),
         Birth_Year_o = if_else(
           the_diff > 0, Birth_Year_1, Birth_Year_2),
         Birth_Year_y = if_else(
           the_diff > 0, Birth_Year_2, Birth_Year_1)) |>
  filter(rel_type != 'Parental' | abs(the_diff) > 10) |>
  select(rel_type, Kinship,
         ends_with('_o'), ends_with('_y')) |>
  arrange(Birth_Year_o, Birth_Year_y)
```

```{r join_all_first_degree_relatives}
ukb_types <- bind_rows(
    ukb_rel |>
      filter(grepl('MZ', dist)) |>
      inner_join(
        ukb_demo |>
          select(eid,
                 Sex_1=Sex,
                 Birth_Year_1=Birth_Year),
        by=c('ID1'='eid')) |>
      inner_join(
        ukb_demo |>
          select(eid,
                 Sex_2=Sex,
                 Birth_Year_2=Birth_Year),
          by=c('ID2'='eid')) |>
      mutate(rel_type = 'Twin') |>
      select(
        rel_type, eid_o = ID1, Sex_o = Sex_1,
        Birth_Year_o = Birth_Year_1,
        eid_y = ID2, Sex_y = Sex_2,
        Birth_Year_y = Birth_Year_2,
        Kinship),
    ukb_first_types) |>
  arrange(Birth_Year_o, Birth_Year_y)
```

```{r save_all_first_degree_relatives}
ukb_types |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_rel_first_type.csv'),
    na = '')
```

## Trios

```{r identify_relative_trios}
ukb_trios <- ukb_types |>
  group_by(eid_y) |>
  filter(rel_type == 'Parental') |>
  filter(n_distinct(Sex_o) > 1)  |>
  ungroup() |>
  arrange(eid_y, Sex_o) |>
  select(-rel_type) |>
  relocate(ends_with('_y'))
```

```{r save_relative_trios}
ukb_trios |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_rel_trios.csv'),
    na = '')
```

## Extended Families

```{r identify_families}
ukb_rel_eids_l <- ukb_rel_ages |>
  group_by(eid_o) |>
  summarise(Birth_Year_o = unique(Birth_Year_o),
            eids = list(eid_y),
            .groups = 'keep') |>
  mutate(eids = list(c(eid_o, unlist(eids)))) |>
  pull(eids)

the_rows <- rep(1:length(ukb_rel_eids_l), lengths(ukb_rel_eids_l)) 
the_cols <- factor(unlist(ukb_rel_eids_l))
ukb_rel_eids_t <- Matrix::sparseMatrix(
  i = the_rows, j = as.integer(the_cols),
  x = TRUE, dimnames = list(NULL, levels(the_cols)))

ukb_rel_eids_c <- Matrix::tcrossprod(ukb_rel_eids_t,
                                     boolArith = TRUE)

ukb_rel_eids_g <- igraph::clusters(
  igraph::graph_from_adjacency_matrix(
    as(ukb_rel_eids_c, "lMatrix")))$membership

ukb_families_l <- tapply(ukb_rel_eids_l,
                   ukb_rel_eids_g, function(x) sort(unique(unlist(x))))

ukb_families <- as_tibble(
  setNames(stack(ukb_families_l)[2:1],
  c('fam_id', 'eid')))
```

```{r save_families}
ukb_families |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_families.csv'),
    na = '')
```

### Supplementary: Hypothetical Example

```{r define_hyp_relatives}
hyp_by <- tribble(
  ~eid, ~Birth_Year,
  'A', 1940,
  'B', 1965,
  'C', 1965,
  'D', 1945,
  'E', 1930,
  'F', 1950,
  'G', 1980,
  'H', 1935
)

hyp_rel <- tribble(
  ~rid, ~eid_1, ~eid_2,
  1, 'A', 'B',
  2, 'B', 'C',
  3, 'C', 'A',
  4, 'C', 'D',
  5, 'E', 'F',
  6, 'E', 'H',
  7, 'F', 'G',
  8, 'G', 'E',
  9, 'G', 'H',
  10, 'H', 'F'
)

hyp_rel_by <- hyp_rel |> 
  inner_join(hyp_by, by = c('eid_1' = 'eid')) |>
  inner_join(hyp_by, by = c('eid_2' = 'eid'), suffix = c('_1', '_2')) |>
  mutate(
    eid_o = if_else(Birth_Year_2 > Birth_Year_1, eid_1, eid_2),
    eid_y = if_else(eid_o == eid_1, eid_2, eid_1),
    Birth_Year_o = if_else(Birth_Year_2 > Birth_Year_1,
                           Birth_Year_1, Birth_Year_2),
    Birth_Year_y = if_else(eid_o == eid_1, Birth_Year_2, Birth_Year_1)) |>
  select(rid, ends_with(c('_o', 'y'))) |>
  arrange(Birth_Year_o, eid_o, Birth_Year_y, eid_y)
```

```{r identify_hyp_fam_step1}
hyp_rel_eids_l <- hyp_rel_by |>
  group_by(eid_o) |>
  summarise(Birth_Year_o = unique(Birth_Year_o),
            eids = list(eid_y),
            .groups = 'keep') |>
  mutate(eids = list(c(eid_o, unlist(eids)))) |>
  pull(eids)
```

```{r identify_hyp_fam_step2}
hyp_rows <- rep(1:length(hyp_rel_eids_l), lengths(hyp_rel_eids_l))
```

```{r identify_hyp_fam_step3}
hyp_cols <- factor(unlist(hyp_rel_eids_l))
```

```{r identify_hyp_fam_step4}
hyp_rel_eids_t <- Matrix::sparseMatrix(
  i = hyp_rows, j = as.integer(hyp_cols),
  x = TRUE, dimnames = list(paste0('G', unique(hyp_rows)),
                            levels(hyp_cols)))
```

```{r identify_hyp_fam_step5}
hyp_rel_eids_c <- Matrix::tcrossprod(hyp_rel_eids_t,
                                     boolArith = TRUE)
```

```{r identify_hyp_fam_step6}
hyp_rel_eids_g <- igraph::clusters(
  igraph::graph_from_adjacency_matrix(
    as(hyp_rel_eids_c, "lMatrix")))$membership
```

```{r identify_hyp_fam_step7}
hyp_families_l <- tapply(
  hyp_rel_eids_l,
  hyp_rel_eids_g, function(x) sort(unique(unlist(x))))
```