---
title: "Figure 2.- Association between rDNA CN and hematological profiles"
---

## Setup

```{r setup}
source('0.0.- Setup.R')
```

```{r read_cn_wb}
ukb_47s_wb <- read_csv(
  file.path(root_fd, 'data', 'cn', 'ukb_47s_wb.csv'),
  col_types = cols())
```

```{r read_demographics}
ukb_demo <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_demo.csv'),
  col_types = cols())
```

```{r read_seq_rel}
ukb_sr <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_sr.csv'),
  col_types = cols())
```

```{r read_covars}
ukb_gwas_covars <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_gwas_covars.csv'),
  col_types = cols())
```

## PHESANT Control

```{r read_phesant_control}
phesant_control <- clean_phesant(read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_with_bc', 'cn_shuffled', 'results-combined.txt'),
  col_types = cols())) |>
  mutate(cat = stringr::str_squish(sub(".*>", "", Path)),
         category = if_else(
           grepl('Blood assay|Medications', Cat3_Title), cat, Cat1_Title)) |>
  select(varName, category, description,
         N = total_n, cases, beta, ci_low = lower, ci_high = upper,
         pvalue) |>
  arrange(pvalue) |>
  mutate(description = gsub('ophill', 'ophil', description))
```

```{r filter_phesant_control}
phesant_control_fdr <- phesant_control |>
  mutate(fdr = p.adjust(pvalue, method = 'fdr')) |>
  filter(fdr < 0.01) |>
  arrange(fdr)
```

## A.- PHESANT CN

### Main.- Both Sequencing Centres

```{r read_phesant_cn}
phesant_cn_with_blood <- clean_phesant(read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_with_bc', 'cn', 'results-combined.txt'),
  col_types = cols())) |>
  mutate(cat = stringr::str_squish(sub(".*>", "", Path)),
         category = if_else(
           grepl('Blood assay|Medications', Cat3_Title), cat, Cat1_Title)) |>
  select(varName, category, description,
         N = total_n, cases, beta, ci_low = lower, ci_high = upper,
         pvalue) |>
  arrange(pvalue) |>
  mutate(description = gsub('ophill', 'ophil', description))
```

```{r filter_phesant_cn}
phesant_cn_with_blood_fdr <- phesant_cn_with_blood |>
  mutate(fdr = p.adjust(pvalue, method = 'fdr')) |>
  filter(fdr < 0.01) |>
  arrange(fdr) |>
  mutate(description = case_when(
    grepl('F06', description) ~ 'F06: Other mental disorders [...]',
    grepl('D73', description) ~ 'D73: Diseases of spleen',
    TRUE ~ description),
    description = gsub('Type of cancer: ICD10: O01.9', 'O01.9:', description),
    description = gsub('Treatment/medication code: s', 'S', description),
    description = gsub('White blood cell \\(leukocyte\\)',
                       'Leukocyte', description),
    description = gsub('Red blood cell \\(erythrocyte\\)', 
                       'Erythrocyte', description),
    description = factor(description) |>
      forcats::fct_inorder() |>
      forcats::fct_rev()) |>
  mutate(category = gsub('d o', 'd\no', category),
         supercat = if_else(!grepl('Blood', category), 'Other', category))
```

```{r plot_phesant_cn_betas}
phesant_cn_with_blood_beta_p <- phesant_cn_with_blood_fdr |>
  mutate(supercat = fct_rev(supercat)) |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.042, 0.042),
                     breaks = c(-0.04, -0.02, 0, 0.02, 0.04)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_sig}
phesant_cn_with_blood_fdr_p <- phesant_cn_with_blood_fdr |>
  mutate(supercat = fct_rev(gsub(' b', '\nb', supercat))) |>
  mutate(lfdr = -log10(fdr)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 41)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_plot}
phesant_cn_with_blood_p <- (phesant_cn_with_blood_beta_p |
  phesant_cn_with_blood_fdr_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_with_blood_p,
  filename = file.path('..', 'figures', '2.A.- PHESANT CN.png'), 
  height = 7, width = 8.5)
```

```{r plot_phesant_cn_betas_bottom}
phesant_cn_with_blood_beta_other_p <- phesant_cn_with_blood_fdr |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.35, 0.35),
                     breaks = c(-0.3, 0, 0.3)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_sig_bottom}
phesant_cn_with_blood_fdr_other_p <- phesant_cn_with_blood_fdr |>
  mutate(lfdr = -log10(fdr)) |>
  mutate(supercat = gsub(' b', '\nb', supercat)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 41)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_betas_bottom_plot}
phesant_cn_with_blood_p <- (phesant_cn_with_blood_beta_other_p |
  phesant_cn_with_blood_fdr_other_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_with_blood_p,
  filename = file.path('..', 'figures', '2.A.- PHESANT CN, Other.png'), 
  height = 7, width = 8.5)
```

### Supplementary.- Sanger

```{r read_phesant_cn_sanger}
phesant_cn_sanger <- clean_phesant(read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_sc', 'cn_sanger', 'results-combined.txt'),
  col_types = cols()), min_recorded = 85007 / 2, min_cases = 100) |>
  mutate(cat = stringr::str_squish(sub(".*>", "", Path)),
         category = if_else(
           grepl('Blood assay|Medications', Cat3_Title), cat, Cat1_Title)) |>
  select(varName, category, description,
         N = total_n, cases, beta, ci_low = lower, ci_high = upper,
         pvalue) |>
  arrange(pvalue) |>
  mutate(description = gsub('ophill', 'ophil', description))
```

```{r filter_phesant_cn_sanger}
phesant_cn_sanger_fdr <- phesant_cn_sanger |>
  mutate(fdr = p.adjust(pvalue, method = 'fdr')) |>
  filter(varName %in% phesant_cn_with_blood_fdr$varName) |>
  mutate(description = case_when(
    grepl('F06', description) ~ 'F06: Other mental disorders [...]',
    grepl('D73', description) ~ 'D73: Diseases of spleen',
    TRUE ~ description),
    description = gsub('Type of cancer: ICD10: O01.9', 'O01.9:', description),
    description = gsub('Treatment/medication code: s', 'S', description),
    description = gsub('White blood cell \\(leukocyte\\)',
                       'Leukocyte', description),
    description = gsub('Red blood cell \\(erythrocyte\\)', 
                       'Erythrocyte', description),
    description = factor(description) |>
      forcats::fct_inorder() |>
      forcats::fct_rev()) |>
  mutate(category = gsub('d o', 'd\no', category),
         supercat = if_else(!grepl('Blood', category), 'Other', category))
``` 

```{r plot_phesant_cn_sanger_betas}
phesant_cn_sanger_beta_p <- phesant_cn_sanger_fdr |>
  mutate(supercat = fct_rev(supercat)) |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.0452, 0.0452),
                     breaks = c(-0.04, -0.02, 0, 0.02, 0.04)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_sanger_sig}
phesant_cn_sanger_fdr_p <- phesant_cn_sanger_fdr |>
  mutate(supercat = fct_rev(gsub(' b', '\nb', supercat))) |>
  mutate(lfdr = -log10(fdr)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 30)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_sanger_plot}
phesant_cn_sanger_p <- (phesant_cn_sanger_beta_p |
  phesant_cn_sanger_fdr_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_sanger_p,
  filename = file.path('..', 'figures_supp', 'S10.- PHESANT CN, Sanger.png'), 
  height = 5.5, width = 8.5)
```

```{r plot_phesant_cn_sanger_betas_bottom}
phesant_cn_sanger_beta_other_p <- phesant_cn_sanger_fdr |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.4, 0.4),
                     breaks = c(-0.3, 0, 0.3)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_sanger_sig_bottom}
phesant_cn_sanger_fdr_other_p <- phesant_cn_sanger_fdr |>
  mutate(lfdr = -log10(fdr)) |>
  mutate(supercat = gsub(' b', '\nb', supercat)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 30)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_sanger_betas_bottom_plot}
phesant_cn_sanger_other_p <- (phesant_cn_sanger_beta_other_p |
  phesant_cn_sanger_fdr_other_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_sanger_other_p,
  filename = file.path('..', 'figures_supp',
                       'S10.- PHESANT CN, Other, Sanger.png'), 
  height = 5.5, width = 8.5)
```

### Supplementary.- deCODE

```{r read_phesant_cn_decode}
phesant_cn_decode <- clean_phesant(read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_sc', 'cn_decode', 'results-combined.txt'),
  col_types = cols()), min_recorded = 72220 / 2, min_cases = 100) |>
  mutate(cat = stringr::str_squish(sub(".*>", "", Path)),
         category = if_else(
           grepl('Blood assay|Medications', Cat3_Title), cat, Cat1_Title)) |>
  select(varName, category, description,
         N = total_n, cases, beta, ci_low = lower, ci_high = upper,
         pvalue) |>
  arrange(pvalue) |>
  mutate(description = gsub('ophill', 'ophil', description))
```

```{r filter_phesant_cn_decode}
phesant_cn_decode_fdr <- phesant_cn_decode |>
  mutate(fdr = p.adjust(pvalue, method = 'fdr')) |>
  filter(varName %in% phesant_cn_with_blood_fdr$varName) |>
  mutate(description = case_when(
    grepl('F06', description) ~ 'F06: Other mental disorders [...]',
    grepl('D73', description) ~ 'D73: Diseases of spleen',
    TRUE ~ description),
    description = gsub('Type of cancer: ICD10: O01.9', 'O01.9:', description),
    description = gsub('Treatment/medication code: s', 'S', description),
    description = gsub('White blood cell \\(leukocyte\\)', 
                       'Leukocyte', description),
    description = gsub('Red blood cell \\(erythrocyte\\)', 
                       'Erythrocyte', description),
    description = factor(description) |>
      forcats::fct_inorder() |>
      forcats::fct_rev()) |>
  mutate(category = gsub('d o', 'd\no', category),
         supercat = if_else(!grepl('Blood', category), 'Other', category))
```

```{r plot_phesant_cn_decode_betas}
phesant_cn_decode_beta_p <- phesant_cn_decode_fdr |>
  mutate(supercat = fct_rev(supercat)) |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.0452, 0.0452),
                     breaks = c(-0.04, -0.02, 0, 0.02, 0.04)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_decode_sig}
phesant_cn_decode_fdr_p <- phesant_cn_decode_fdr |>
  mutate(supercat = fct_rev(gsub(' b', '\nb', supercat))) |>
  mutate(lfdr = -log10(fdr)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 30)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_decode_plot}
phesant_cn_decode_p <- (phesant_cn_decode_beta_p |
  phesant_cn_decode_fdr_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_decode_p,
  filename = file.path('..', 'figures_supp',
                       'S10.- PHESANT CN, deCODE.png'), 
  height = 5.5, width = 8.5)
```

```{r plot_phesant_cn_decode_betas_bottom}
phesant_cn_decode_beta_other_p <- phesant_cn_decode_fdr |>
  ggplot(aes(x = beta, xmin = ci_low, xmax = ci_high, y = description)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(shape = category, fill = category),
             alpha = 0.8, size = 2) +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  labs(y = '', x = 'Standardised Beta, 95% CI') +
  scale_x_continuous(limits = c(-0.4, 0.4),
                     breaks = c(-0.3, 0, 0.3)) +
  scale_fill_manual('Phenotype category',
                    values = c('steelblue3', 'firebrick3', 
                               'gold3', 'olivedrab3')) +
  scale_shape_manual('Phenotype category',
                     values = c(24, 23, 22, 21)) +
  theme_pub() +
  theme(legend.position = 'none',
        legend.title = element_text(face = 'bold',
                                    size = 12),
        legend.text = element_text(size = 11),
        legend.spacing.y = unit(0.3, 'lines'),
        legend.box.background = element_rect(linewidth = 0.4),
        strip.text = element_blank()) 
```

```{r plot_phesant_cn_decode_sig_bottom}
phesant_cn_decode_fdr_other_p <- phesant_cn_decode_fdr |>
  mutate(lfdr = -log10(fdr)) |>
  mutate(supercat = gsub(' b', '\nb', supercat)) |>
  ggplot(aes(x = lfdr, y = description)) +
  geom_col(fill = 'grey90', colour = 'grey20', linewidth = 0.75,
           width = 0.8) +
  geom_vline(xintercept = -log10(0.01), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(supercat), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(0, 30)) +
  labs(y = '', x = bquote(bold(-log[10](FDR)))) +
  theme_pub() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y = element_text(size = 10))
```

```{r save_phesant_cn_decode_bottom_plot}
phesant_cn_decode_other_p <- (phesant_cn_decode_beta_other_p |
  phesant_cn_decode_fdr_other_p) +
  plot_layout(widths = c(0.65, 0.35))

ggsave(
  plot = phesant_cn_decode_other_p,
  filename = file.path('..', 'figures_supp',
                       'S10.- PHESANT CN, Other, deCODE.png'), 
  height = 5.5, width = 8.5)
```

## B.- Association with Blood Counts and Ratios

```{r read_blood_counts}
ukb_blood_counts <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_blood_counts.csv'),
  col_types = cols())
```

```{r construct_blood_counts_covars}
ukb_covars_bc <- ukb_gwas_covars |>
  inner_join(ukb_blood_counts, by = 'eid') |>
  select(-Height, -starts_with('Blood'))
```

### Supplementary.- Multivariate Blood Counts

```{r fit_blood_counts}
ukb_bc_sig_scale <- broom::tidy(
  lm(CN ~ . , ukb_covars_bc |>
       mutate(across(where(is.numeric), scale)) |>
       select(-eid)), 
  conf.int = TRUE) |>
  mutate(lpval = -log10(p.value),
         scoeff = sign(estimate)) |>
  filter(grepl('#', term)) |>
  arrange(lpval) |>
  mutate(term = forcats::fct_inorder(gsub('`', '', term)))
```

```{r plot_blood_counts_betas}
ukb_bc_beta_p <- ukb_bc_sig_scale |>
  ggplot(aes(y = term, x = estimate,
             xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(fill = as.factor(scoeff)),
                  alpha = 0.8, colour = 'grey20',size = 2,
             shape = 23) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  scale_x_continuous(limits = c(-0.06, 0.06),
                     breaks = c(-0.05, 0, 0.05)) +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  theme_pub() +
  theme(legend.position = 'none',
        strip.text = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 15))
```

```{r save_blood_counts_betas_plot}
ggsave(
  plot = ukb_bc_beta_p,
  filename = file.path('..', 'figures_supp',
                       'Blood Counts, Association Betas.png'), 
  height = 6, width = 4)
```

```{r plot_blood_counts_sig}
ukb_bc_sig_p <- ukb_bc_sig_scale |> 
  ggplot(aes(y = term, x = lpval, fill = as.factor(scoeff))) +
  geom_col(alpha = 0.8, colour = 'grey20', linewidth = 0.75) +
  geom_vline(xintercept = -log10(0.05), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  theme_pub() +
  theme(legend.position = 'none',
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_blank())
```

```{r save_blood_counts_sig_plot}
ggsave(
  plot = ukb_bc_sig_p,
  filename = file.path('..', 'figures_supp',
                       'Blood Ratios, Association p-values.png'), 
  height = 6, width = 4)
```

```{r save_blood_counts_association_plot}
ukb_bc_p <- (ukb_bc_beta_p +
   theme(
     plot.margin = margin(r = -100)) |
  (ukb_bc_sig_p + theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(l = -100)))) +
  plot_layout(widths = c(0.55, 0.45))

ggsave(
  plot = ukb_bc_p,
  filename = file.path('..', 'figures_supp',
                       'Blood Ratios, Association.png'), 
  height = 6, width = 7.5)
```

### Main.- Blood Composition Ratios

```{r define_group_labels}
group_labels <- c('Low', 'Mid-Low', 'Mid', 'Mid-High', 'High')
```

```{r construct_blood_ratios_covars}
ukb_covars_br <- ukb_covars_bc |>
  filter(`Monocyte #` > 0, !is.na(`Platelet #`)) |>
  mutate(NLR = `Neutrophil #` / `Lymphocyte #`,
         NLR_group = cut_number(NLR, n = 5, labels = group_labels),
         PLR = `Platelet #` / `Lymphocyte #`,
         PLR_group = cut_number(PLR, n = 5, labels = group_labels),
         LMR = `Lymphocyte #` / `Monocyte #`,
         LMR_group = cut_number(LMR, n = 5, labels = group_labels),
         SII = (`Neutrophil #` *`Platelet #`) / `Lymphocyte #`,
         SII_group = cut_number(SII, n = 5, labels = group_labels))
```

```{r save_blood_ratios_covars}
ukb_covars_br |> 
  select(eid, NLR, PLR, LMR, SII) |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_blood_ratios.csv'),
    na = '')
```

```{r fit_blood_ratios}
ukb_br_sig_scale <- bind_rows(
  lapply(X = c('NLR', 'PLR', 'LMR', 'SII'),
         FUN = function(x) fit_ratio_scale(ukb_covars_br, x))) |>
  arrange(lpval) |>
  mutate(term = forcats::fct_inorder(term))
```

```{r plot_blood_ratios_betas}
ukb_br_beta_p <- ukb_br_sig_scale |>
  ggplot(aes(y = term, x = estimate,
             xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_crossbar(fill = 'grey90', colour = 'grey60',
                linewidth = 0.5, fatten = 1, alpha = 0.5) +
  geom_point(aes(fill = as.factor(scoeff)),
                  alpha = 0.8, colour = 'grey20',size = 2,
             shape = 23) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  scale_x_continuous(limits = c(-0.06, 0.06),
                     breaks = c(-0.05, 0, 0.05)) +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  theme_pub() +
  theme(legend.position = 'none',
        strip.text = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 15))
```

```{r save_blood_ratios_betas_plot}
ggsave(
  plot = ukb_br_beta_p,
  filename = file.path('..', 'figures',
                       '2.B.- Blood Ratios, Association Betas.png'), 
  height = 2, width = 4)
```

```{r plot_blood_ratios_sig}
ukb_br_sig_p <- ukb_br_sig_scale |> 
  ggplot(aes(y = term, x = lpval, fill = as.factor(scoeff))) +
  geom_col(alpha = 0.8, colour = 'grey20', linewidth = 0.75) +
  geom_vline(xintercept = -log10(0.05), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  theme_pub() +
  theme(legend.position = 'none',
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_blank())
```

```{r save_blood_ratios_sig_plot}
ggsave(
  plot = ukb_br_sig_p,
  filename = file.path('..', 'figures', 
                       '2.B.- Blood Ratios, Association p-values.png'), 
  height = 2.5, width = 4)
```

```{r save_blood_ratios_association_plot}
ukb_br_p <- (ukb_br_beta_p +
   theme(
     plot.margin = margin(r = -100)) |
  (ukb_br_sig_p + theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(l = -100)))) +
  plot_layout(widths = c(0.6, 0.4))

ggsave(
  plot = ukb_br_p,
  filename = file.path('..', 'figures', 
                       '2.B.- Blood Ratios, Association.png'), 
  height = 2.2, width = 7.2)
```

## Mendelian Randomisation

```{r read_cn_gwas_results}
bolt_lmm_cn_raw <- read_tsv(
  file.path(root_fd, 'data', 'gwas_imp', 'cn_msnps.tsv'),
  col_types = cols())
```

```{r format_cn_gwas_results}
cn_ass_format <- bolt_lmm_cn_raw %>%
  mutate(Phenotype = '18S Ratio') %>%
  TwoSampleMR::format_data(
    type = 'outcome',
    snp_col = 'SNP',
    beta_col = "BETA",
    se_col = "SE",
    effect_allele_col = "ALLELE1",
    other_allele_col = "ALLELE0",
    eaf_col = "A1FREQ",
    pval_col = "P_BOLT_LMM_INF",
    )
```

```{r read_gwas_catalog}
gwas_catalog <- read_tsv(
  file.path(
    root_fd, '..', 'data', 'gwas_catalog',
    'gwas_catalog_v1.0.2-associations_e108_r2023-01-14.tsv'),
  col_types = cols())
```

#### Neutrophil counts

```{r read_neutro_counts_gwas_results}
nc_associations <- gwas_catalog |> 
  filter(grepl('neutrophil count', MAPPED_TRAIT),
         `DISEASE/TRAIT` == 'Neutrophil count',
         grepl('European|British', `INITIAL SAMPLE SIZE`),
         !grepl('NR', `INITIAL SAMPLE SIZE`))
```

```{r format_neutro_counts_gwas_results}
suppressWarnings(
  nc_ass_format <- nc_associations |>
    mutate(se = TwoSampleMR::get_se(`OR or BETA`, `P-VALUE`)) |>
    select(SNP=starts_with('STRONGEST'),
           beta=`OR or BETA`,
           pval=`P-VALUE`, 
           se,
           eaf = ends_with('FREQUENCY')) |>
    separate(SNP, into = c('SNP', 'effect_allele'), 
             sep = '-', extra = 'warn') |>
    group_by(SNP) |>
    filter(n_distinct(effect_allele) == 1) |>
    group_by(SNP, effect_allele) |>
    filter(pval == min(pval)) |>
    filter(beta == max(beta)) |>
    mutate(pheno = 'Neutrophil count') |>
    TwoSampleMR::format_data(phenotype_col = 'pheno',
                             type = 'exposure')) 
```

```{r clump_neutro_counts_gwas_results}
nc_ass_clump <- TwoSampleMR::clump_data(nc_ass_format)
```

```{r harmonise_neutro_counts_gwas_results}
nc_cn_harm_clump <- TwoSampleMR::harmonise_data(
    exposure_dat = nc_ass_clump, 
    outcome_dat = cn_ass_format)
```

```{r calculate_mr_neutro_counts}
(nc_cn_mc_clump <- TwoSampleMR::mr(nc_cn_harm_clump,
                                   method = 'mr_ivw'))
```

#### Neutrophil percentage

```{r read_neutro_perc_gwas_results}
np_associations <- gwas_catalog |> 
  filter(grepl('neutrophil', MAPPED_TRAIT),
         grepl('Neutrophil percentage of white cells', `DISEASE/TRAIT`),
         grepl('European|British', `INITIAL SAMPLE SIZE`),
         !grepl('NR', `INITIAL SAMPLE SIZE`))
```

```{r format_neutro_perc_gwas_results}
suppressWarnings(
  np_ass_format <- np_associations |>
    mutate(se = TwoSampleMR::get_se(`OR or BETA`, `P-VALUE`)) |>
    select(SNP=starts_with('STRONGEST'),
           beta=`OR or BETA`,
           pval=`P-VALUE`, 
           se,
           eaf = ends_with('FREQUENCY')) |>
    separate(SNP, into = c('SNP', 'effect_allele'), 
             sep = '-', extra = 'warn') |>
    group_by(SNP) |>
    filter(n_distinct(effect_allele) == 1) |>
    group_by(SNP, effect_allele) |>
    filter(pval == min(pval)) |>
    filter(beta == max(beta)) |>
    mutate(pheno = 'Neutrophil percentage') |>
    TwoSampleMR::format_data(phenotype_col = 'pheno',
                             type = 'exposure')) 
```

```{r clump_neutro_perc_gwas_results}
np_ass_clump <- TwoSampleMR::clump_data(np_ass_format)
```

```{r harmonise_neutro_perc_gwas_results}
np_cn_harm_clump <- TwoSampleMR::harmonise_data(
    exposure_dat = np_ass_clump, 
    outcome_dat = cn_ass_format)
```

```{r calculate_mr_neutro_perc}
(np_cn_mc_clump <- TwoSampleMR::mr(np_cn_harm_clump,
                                   method = c('mr_ivw')))
```

#### Platelet counts

```{r read_plat_counts_gwas_results}
pc_associations <- gwas_catalog |> 
  filter(grepl('platelet count', MAPPED_TRAIT),
         `DISEASE/TRAIT` == 'Platelet count',
         grepl('European|British', `INITIAL SAMPLE SIZE`),
         !grepl('NR', `INITIAL SAMPLE SIZE`))
```

```{r format_plat_counts_gwas_results}
suppressWarnings(
  pc_ass_format <- pc_associations |>
    mutate(se = TwoSampleMR::get_se(`OR or BETA`, `P-VALUE`)) |>
    select(SNP=starts_with('STRONGEST'),
           beta=`OR or BETA`,
           pval=`P-VALUE`, 
           se,
           eaf = ends_with('FREQUENCY')) |>
    separate(SNP, into = c('SNP', 'effect_allele'), 
             sep = '-', extra = 'warn') |>
    group_by(SNP) |>
    filter(n_distinct(effect_allele) == 1) |>
    group_by(SNP, effect_allele) |>
    filter(pval == min(pval)) |>
    filter(beta == max(beta)) |>
    mutate(pheno = 'Platelet count') |>
    TwoSampleMR::format_data(phenotype_col = 'pheno',
                             type = 'exposure')) 
```

```{r clump_plat_counts_gwas_results}
pc_ass_clump <- TwoSampleMR::clump_data(pc_ass_format)
```

```{r harmonise_plat_counts_gwas_results}
pc_cn_harm_clump <- TwoSampleMR::harmonise_data(
    exposure_dat = pc_ass_clump, 
    outcome_dat = cn_ass_format)
```

```{r calculate_mr_plat_counts}
(pc_cn_mc_clump <- TwoSampleMR::mr(pc_cn_harm_clump, method = c('mr_ivw')))
```

#### Lymphocyte percentage

```{r read_lymph_perc_gwas_results}
lp_associations <- gwas_catalog |> 
  filter(grepl('lymphocyte', MAPPED_TRAIT),
         `DISEASE/TRAIT` == 'Lymphocyte percentage of white cells',
         grepl('European|British', `INITIAL SAMPLE SIZE`),
         !grepl('NR', `INITIAL SAMPLE SIZE`))
```

```{r format_lymph_perc_gwas_results}
suppressWarnings(
  lp_ass_format <- lp_associations |>
    mutate(se = TwoSampleMR::get_se(`OR or BETA`, `P-VALUE`)) |>
    select(SNP=starts_with('STRONGEST'),
           beta=`OR or BETA`,
           pval=`P-VALUE`, 
           se,
           eaf = ends_with('FREQUENCY')) |>
    separate(SNP, into = c('SNP', 'effect_allele'), 
             sep = '-', extra = 'warn') |>
    group_by(SNP) |>
    filter(n_distinct(effect_allele) == 1) |>
    group_by(SNP, effect_allele) |>
    filter(pval == min(pval)) |>
    filter(beta == max(beta)) |>
    mutate(pheno = 'Lymphocyte percentage') |>
    TwoSampleMR::format_data(phenotype_col = 'pheno',
                             type = 'exposure')) 
```

```{r clump_lymph_perc_gwas_results}
lp_ass_clump <- TwoSampleMR::clump_data(lp_ass_format)
```

```{r harmonise_lymph_perc_gwas_results}
lp_cn_harm_clump <- TwoSampleMR::harmonise_data(
    exposure_dat = lp_ass_clump, 
    outcome_dat = cn_ass_format)
```

```{r calculate_mr_lymph_perc}
(lp_cn_mc_clump <- TwoSampleMR::mr(lp_cn_harm_clump, method = c('mr_ivw')))
```

### Summary

```{r join_mr_results}
bind_rows(
  lp_cn_mc_clump,
  nc_cn_mc_clump,
  np_cn_mc_clump,
  pc_cn_mc_clump) |>
  select(-starts_with('id')) |>
  write_csv('../tables/MR Results.csv',
            na = '')
```

```{r save_mr_sources}
bind_rows(
  lp_associations,
  nc_associations,
  np_associations,
  pc_associations) |>
  select(`DISEASE/TRAIT`, MAPPED_TRAIT, PUBMEDID,
         `FIRST AUTHOR`, JOURNAL, `STUDY ACCESSION`)  |>
  distinct() |>
  write_csv('../tables/Source of MR associations.csv',
            na = '')
```

## D.- Known influences on NLR

```{r read_nlr}
ukb_nlr <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_nlr.csv'),
  col_types = cols()) |> 
  select(eid, NLR)
```

```{r read_infl_nlr}
ukb_infl_nlr <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_infl_nlr.csv'),
  col_types = cols())
infl_nlr_vars <- ukb_infl_nlr |> select(-eid) |> colnames()
```

```{r read_infl_nlr_cov}
ukb_demo <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_demo.csv'),
  col_types = cols())
ukb_pcs <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_pc.csv'),
  col_types = cols())
ukb_atl <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_adj_tl.csv'),
  col_types = cols())
ukb_ac <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_ac.csv'),
  col_types = cols())
```

```{r construct_infl_nlr_covars}
ukb_nlr_pheno <- ukb_nlr |>
  inner_join(ukb_demo, by = 'eid') |>
  inner_join(ukb_pcs, by = 'eid') |>
  inner_join(ukb_atl, by = 'eid') |>
  inner_join(ukb_ac, by = 'eid') |>
  mutate(`Age Sq` = Age * Age,
         Eth_Back = as.factor(Eth_Back)) |>
  select(eid, Sex, Eth_Back,
         starts_with('Age'),
         contains('Centre'),
         paste0('PC', 1:10),
         starts_with('Adj'),
         NLR) 
```

```{r fit_infl_nlr_age_sex_function}
fit_infl_nlr_age_sex <- function(pop = 'NLR - UKB') {
  
  if(pop == 'NLR - UKB') { 
    df <- ukb_nlr_pheno
  } else if (pop == 'NLR - WB') {
    df <- (ukb_nlr_pheno |>
      filter(Eth_Back == '1001') |>
      select(-Eth_Back))
  } else {
    df <- (ukb_nlr_pheno |>
             filter(eid %in% ukb_gwas_covars$eid) |>
             select(-Eth_Back))
  }
  
  broom::tidy(
    lm(NLR ~ . , df |> 
         mutate(across(where(is.numeric), scale)) |>
         select(-eid)),
    conf.int = TRUE) |> 
    mutate(term = gsub('`', '', term),
           lpval = -log10(p.value),
           scoeff = sign(estimate),
           Population = pop) |>
    filter(grepl('Age', term) | grepl('Sex', term)) |>
    relocate(Population)
}
```

```{r fit_infl_nlr_function}
fit_infl_nlr_var <- function(pop = 'NLR - UKB',
                             the_var = 'Cancer') {
  
  aux <- ukb_nlr_pheno |>
      inner_join(ukb_infl_nlr |> select(eid, all_of(the_var)),
                 by = 'eid')
  
  if(pop == 'NLR - UKB') { 
    df <- aux
  } else if (pop == 'NLR - WB') {
    df <- (aux |>
      filter(Eth_Back == '1001') |>
      select(-Eth_Back))
  } else {
    df <- (aux |>
             filter(eid %in% ukb_gwas_covars$eid) |>
             select(-Eth_Back))
  }
  
  broom::tidy(
    lm(NLR ~ . , df |> 
         mutate(across(where(is.numeric), scale)) |>
         select(-eid)),
    conf.int = TRUE) |> 
    mutate(term = gsub('`', '', term),
           lpval = -log10(p.value),
           scoeff = sign(estimate),
           Population = pop) |>
    filter(grepl(the_var, term)) |>
    mutate(term = gsub('Yes', '', term)) |> 
    relocate(Population)
}
```

```{r define_populations_analysis}
populations <- paste('NLR', c('UKB', 'WB', 'WB WGS'),
                     sep = ' - ')
```

```{r fit_infl_nlr}
ukb_nlr_ass <- bind_rows(
  bind_rows(
    lapply(
      X = populations,
      FUN = function(x) fit_infl_nlr_age_sex(pop = x))),
  bind_rows(
    lapply(
      X = populations,
      FUN = function(x) lapply(
        X = infl_nlr_vars,
        FUN = function(y) fit_infl_nlr_var(pop = x,
                                           the_var = y))))) |>
  arrange(Population, term)
```

```{r fit_cn_age_sex_function}
fit_cn_age_sex <- function() {
  
  df <- ukb_gwas_covars |>
    select(-starts_with('Blood'), -Height)
  
  broom::tidy(
    lm(CN ~ . , df |>
         mutate(across(where(is.numeric), scale)) |>
         select(-eid)),
    conf.int = TRUE) |> 
    mutate(term = gsub('`', '', term),
           lpval = -log10(p.value),
           scoeff = sign(estimate),
           Population = '18S Ratio - WB WGS') |>
    filter(grepl('Age', term) | grepl('Sex', term)) |>
    relocate(Population)

}
```

```{r fit_cn_infl_function}
fit_cn_var <- function(the_var = 'Cancer') {
  
  df <- ukb_gwas_covars |>
    select(-starts_with('Blood'), -Height) |>
      inner_join(ukb_infl_nlr |> select(eid, all_of(the_var)),
                 by = 'eid')
  
    broom::tidy(
    lm(CN ~ . , df |>
         mutate(across(where(is.numeric), scale)) |>
         select(-eid)),
    conf.int = TRUE) |> 
    mutate(term = gsub('`', '', term),
           lpval = -log10(p.value),
           scoeff = sign(estimate),
           Population = '18S Ratio - WB WGS') |>
    filter(grepl(the_var, term)) |>
    mutate(term = gsub('Yes', '', term)) |> 
    relocate(Population)
    
}
```

```{r fit_cn_infl}
ukb_cn_ass <- bind_rows(
  fit_cn_age_sex(),
  bind_rows(lapply(
        X = infl_nlr_vars,
        FUN = function(y) fit_cn_var(the_var = y))))
```

```{r join_infl_associations}
infl_nlr_ass <- bind_rows(
  ukb_nlr_ass,
  ukb_cn_ass) |>
  rename(Analysis = Population, Variable = term) |>
  mutate(Analysis = case_when(
    grepl('18S', Analysis) ~ '18S Ratio in WB with WGS',
    grepl('WGS', Analysis) ~ 'NLR in WB with WGS',
    grepl('WB', Analysis) ~ 'NLR in all WB',
    TRUE ~ 'NLR in all UKB',
  )) |>
  mutate(Analysis = fct_relevel(Analysis,
                                '18S Ratio in WB with WGS',
                                after = 3L)) |>
  mutate(Variable = case_when(
    grepl('Sq', Variable) ~ 'Age Squared',
    grepl('Cancer', Variable) ~ 'Cancer',
    grepl('HIV', Variable) ~  'HIV',
    grepl('T2D', Variable) ~  'Type-2 Diabetes',
    grepl('Sex', Variable) ~ 'Sex',
    TRUE ~  'Age'))
```

#### Full plot for supplementary

```{r plot_full_infl_associations}
infl_nlr_sig_full_p <- infl_nlr_ass |>
  mutate(Variable = fct_relevel(Variable, 'Sex')) |> 
  ggplot(aes(x = Variable, y = lpval, 
             fill = Analysis)) +
  geom_col(position = position_dodge(width = 1.05),
           alpha = 0.9, size = 0.75, colour = 'black') +
  geom_hline(yintercept = -log10(0.05),
             colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(cols = vars(Variable), 
             scales = 'free_x') +
  scale_colour_brewer(palette = 'Greys') +
  scale_fill_grey(start = 1, end = 0.1) +
  scale_y_cut(breaks=c(10), space = 0.000001) +
  expand_limits(y = c(-0.5, 310)) +
  labs(x = '',  y = bquote(bold(-log[10](p-value)))) +
  theme_pub() +
  theme(
    legend.position = 'bottom',
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_text(size = 14, face = 'bold'),
    legend.text = element_text(size = 12))
```

```{r save_full_infl_associations_plot}
ggsave(
  plot = infl_nlr_sig_full_p,
  filename = file.path('..', 'figures_supp', 
                       'SXX.- Influences on NLR, Extended.png'), 
  height = 6, width = 12)
```

#### Reduced plot for main

```{r plot_reduced_infl_associations}
infl_nlr_sig_reduced_p <- infl_nlr_ass |>
  filter(!grepl('Squared', Variable),
         !grepl('all UKB', Analysis)) |>
  mutate(
    Variable = gsub('ype-2 Diabetes', '2D', Variable),
    Variable = fct_relevel(Variable, 'Sex')) |> 
  ggplot(aes(x = Variable, y = lpval, 
             fill = Analysis)) +
  geom_col(position = position_dodge(width = 1.05),
           alpha = 0.9, linewidth = 0.75, colour = 'black') +
  geom_hline(yintercept = -log10(0.05),
             colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(cols = vars(Variable), 
             scales = 'free_x') +
  scale_colour_brewer(palette = 'Greys') +
  scale_fill_grey(start = 0.85, end = 0.1) +
  scale_y_cut(breaks=c(10.5), space = 0.0000001) +
  expand_limits(y = c(-0.5, 310)) +
  labs(x = '',  y = bquote(bold(-log[10](p-value)))) +
  theme_pub() +
  theme(
    legend.position = 'bottom',
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_text(size = 14, face = 'bold'),
    legend.text = element_text(size = 12))
```

```{r save_reduced_infl_associations_p}
ggsave(
  plot = infl_nlr_sig_reduced_p,
  filename = file.path('..', 'figures', 
                       '2.D.- Influences on NLR, Reduced.png'), 
  height = 3.75, width = 9)
```

## Temporal Behaviour

```{r read_blood_extr_timing}
ukb_datetime <- read_csv(
  file.path(root_fd, 'data', 'releases', 'pheno_9.csv'),
  col_types = cols()) |>
  select(eid, datetime = `3166-0.0`) |>
  mutate(year = as.factor(lubridate::year(datetime)),
         month = as.factor(lubridate::month(datetime)),
         hour = as.factor(lubridate::hour(datetime)),
         date = lubridate::date(datetime)) |>
  filter(!is.na(datetime)) |>
  filter(as.numeric(as.character(year)) > 2006) |>
  mutate(delay = as.numeric(date - min(date)))
```

#### NLR

```{r combine_nlr_blood_extr_timing}
ukb_nlr_date <- ukb_datetime |>
    inner_join(ukb_nlr |>
      filter(!is.na(NLR)), by = 'eid')
```

```{r fit_nlr_delay}
broom::tidy(lm.beta::lm.beta(
  lm(NLR ~ delay, ukb_nlr_date)))
```

#### Neutrophil Counts

```{r combine_blood_counts_blood_extr_timing}
ukb_bc_delay <- ukb_blood_counts |>
  inner_join(ukb_datetime, by = 'eid')
```

```{r fit_neutro_counts_delay}
broom::tidy(lm.beta::lm.beta(
  lm(`Neutrophil #` ~ delay, ukb_bc_delay)))
```

#### CN

```{r combine_timing_covars}
ukb_covars_date <- ukb_datetime |>
  select(eid, delay) |>
  inner_join(ukb_gwas_covars |> select(-Height), by = 'eid')
```

```{r fit_cn_timing}
broom::tidy(lm.beta::lm.beta(lm(CN ~ .,
           ukb_covars_date |> 
             select(-eid, -starts_with('Blood'))))) |> 
  filter(term == 'delay')
```

```{r combine_cn_blood_counts_timing}
ukb_cn_bc_date <- ukb_datetime |>
  select(eid, delay) |>
  inner_join(ukb_gwas_covars |> select(-Height, -starts_with('Blood')),
             by = 'eid') |>
  inner_join(ukb_blood_counts |>
               filter(!is.na(`Neutrophil #`),
                      !is.na(`Lymphocyte #`),
                      !is.na(`Platelet #`)), by = 'eid')
```

```{r fit_cn_blood_counts_timing}
broom::tidy(lm.beta::lm.beta(lm(CN ~ .,
           ukb_cn_bc_date |> select(-eid)))) |> 
  filter(grepl('#', term))
```

## E.- Seasonal Patterns

```{r slice_nlr_month}
ukb_nlr_month <- ukb_nlr_date |>
    group_by(month) |>
    summarise(value = mean(NLR),
              SD = sd(NLR),
              N = n_distinct(eid),
              margin = qt(ci, df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
    mutate(variable = 'NLR')
```

```{r slice_cn_month}
ukb_cn_adj_month <- ukb_datetime |>
  inner_join(ukb_adj_47s, by = 'eid') |>
  group_by( month) |>
  summarise(value = mean(adj_CN),
            SD = sd(adj_CN),
            N = n_distinct(eid),
            margin = qt(ci,df=N-1)*SD/sqrt(N),
            value_low = value - margin,
            value_high = value + margin,
            .groups = 'drop') |>
  mutate(variable = 'SC-Adjusted 18S Ratio')
```

```{r define_month_colours}
month_colours <- colorRampPalette(
    c('navy', 'steelblue4', 'forestgreen',
            'gold1', 'gold4', 'navy'))(14)[2:13]
```

```{r plot_nlr_month}
monthly_nlr_p <- ukb_nlr_month |>
  ggplot(aes(x = month, y = value, fill = month, colour = month)) +
  geom_crossbar(aes(ymin = value_low, ymax = value_high),
                alpha = 0.4, linewidth = 0.75) +
  facet_grid(rows = vars(variable)) +
  scale_fill_manual(values = month_colours) +
  scale_colour_manual(values = month_colours) +
  scale_y_continuous(
    name = NULL,
    limits = c(2.25, 2.5),
    breaks = c(2.3, 2.4, 2.5)) + 
  theme_pub() +
  theme(
    legend.position = 'none',
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = margin(b = -200))
```

```{r plot_cn_month}
monthly_cn_adj_p <- ukb_cn_adj_month |> 
  arrange(month) |>
  mutate(
    month_n = lubridate::month(as.numeric(month), label=TRUE),
    month_n = forcats::fct_inorder(glue("{month_n}<br>
         <span style='font-size:11pt'>[{N}]</span>"))) |>
  ggplot(aes(x = month_n, y = value, fill = month, colour = month)) +
  geom_crossbar(aes(ymin = value_low, ymax = value_high),
                alpha = 0.4, size = 0.75) +
  facet_grid(rows = vars(variable)) +
  labs(x = 'Month of Blood Extraction') +
  scale_fill_manual(values = month_colours) +
  scale_colour_manual(values = month_colours) +
  scale_y_continuous(
    name = NULL,
    limits = c(1.59e-4, 1.64e-4), 
    breaks = c(1.6e-4, 1.62e-4, 1.64e-4), 
    labels = fancy_scientific) +  
  scale_x_discrete(guide = guide_axis(n.dodge = 1)) +
  theme_pub() +
  theme(
    legend.position = 'none',
    axis.text.x = element_markdown(),
    plot.margin = margin(b = -200, t = -200))
```

```{r save_month_plot}
y_lab <- ggplot() + 
  annotate(geom = "text", x = 1, y = 1, size = 5,
           label = bquote(bold("Trait Mean (" %+-% "95% CI)")),
           angle = 90) +
  coord_cartesian(clip = "off") +
  theme_void()

monthly_p <- (y_lab | (monthly_nlr_p / monthly_cn_adj_p)) +
  plot_layout(widths = c(0.03, 1))

ggsave(
  plot = monthly_p,
  filename = file.path('..', 'figures', '2.E.- Seasonal Patterns.png'), 
  height = 5, width = 10)
```

## F.- Hourly Patterns

```{r slice_nlr_hour}
ukb_nlr_hour <- ukb_datetime |>
    inner_join(ukb_nlr, by = 'eid') |>
    group_by(hour) |>
    filter(n_distinct(eid) > 1000) |>
    summarise(value = mean(NLR),
              SD = sd(NLR),
              N = n_distinct(eid),
              margin = qt(ci,df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
    mutate(variable = 'NLR')
```

```{r slice_cn_hour}
ukb_cn_adj_hour <- ukb_datetime |>
    inner_join(ukb_adj_47s, by = 'eid') |>
    group_by(hour) |>
    filter(n_distinct(eid) > 100) |>
    summarise(value = mean(adj_CN),
              SD = sd(adj_CN),
              N = n_distinct(eid),
              margin = qt(ci,df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
  mutate(variable = 'SC-Adjusted 18S Ratio')
```

```{r define_hour_colours}
hour_colours <- colorRampPalette(
    c('steelblue4',
            'gold1', 'salmon4', 'navy'))(14)[1:13]
```

```{r plot_nlr_hour}
hourly_nlr_p <- ukb_nlr_hour |>
  ggplot(aes(x = hour, y = value, fill = hour, colour = hour)) +
  geom_crossbar(aes(ymin = value_low, ymax = value_high),
                alpha = 0.4, size = 0.75) +
  scale_fill_manual(values = hour_colours) +
  scale_colour_manual(values = hour_colours) +
  facet_grid(rows = vars(variable)) +
  scale_y_continuous(
    name = NULL,
    limits = c(1.8, 2.6),
    breaks = c(1.9, 2.2, 2.5)) + 
  theme_pub() +
  theme(
    legend.position = 'none',
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = margin(l = 2, b = -200))
```

```{r plot_cn_hour}
hourly_cn_adj_p <- ukb_cn_adj_hour |> 
  arrange(hour) |>
  mutate(hour_n = forcats::fct_inorder(
    glue("{hour}h<br><span style='font-size:11.5pt'>[{N}]</span>"))) |>
  ggplot(aes(x = hour_n, y = value, fill = hour, colour = hour)) +
  geom_crossbar(aes(ymin = value_low, ymax = value_high),
                alpha = 0.4, size = 0.75) +
  labs(x = 'Hour of Blood Extraction') +
  scale_fill_manual(values = hour_colours) +
  scale_colour_manual(values = hour_colours) +
  facet_grid(rows = vars(variable)) +
  scale_y_continuous(
    name = NULL,
    limits = c(1.5e-4, 1.7e-4), 
    breaks = c(1.5e-4, 1.6e-4, 1.7e-4), 
    labels = fancy_scientific) + 
  theme_pub() +
  theme(
    axis.text.x = element_markdown(),
    legend.position = 'none',
    plot.margin = margin(l = 5, b = -200, t = -200))
```

```{r save_hour_plot}
y_lab <- ggplot() + 
  annotate(geom = "text", x = 1, y = 1, size = 5,
           label = bquote(bold("Trait Mean (" %+-% "95% CI)")),
           angle = 90) +
  coord_cartesian(clip = "off") +
  theme_void()

hourly_p <- (y_lab | (hourly_nlr_p / hourly_cn_adj_p)) +
  plot_layout(widths = c(0.03, 1))

ggsave(
  plot = hourly_p,
  filename = file.path('..', 'figures', '2.F.- Circadian Patterns.png'), 
  height = 5, width = 10)
```

### Drift

```{r construct_delay_covars}
ukb_delay_covars <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_delay.csv'),
  col_types = cols()) %>%
  mutate(Drift = as.numeric(`Acquisition Time` - min(`Acquisition Time`))) |>
  inner_join(ukb_gwas_covars, by = 'eid') |>
  select(eid, Sex, contains('Age'), contains('Centre'), starts_with('PC'),
         CN, `Adj TL`, Drift, Delay)
```

```{r fit_cn_delay}
broom::tidy(lm(CN ~ ., data = ukb_delay_covars |>
             select(-eid))) |> tail()
```

```{r fit_nlr_delay}
broom::tidy(lm(NLR ~ ., data = ukb_delay_covars |>
             select(-CN) |>
             inner_join(ukb_nlr, by = 'eid'))) |> 
  tail()
```

## G.- Staircase plots

```{r read_adj_cn}
ukb_adj_47s <- read_csv(
  file.path(root_fd, 'data', 'cn', 'ukb_adj_cn_wb.csv'),
  col_types = cols())
```

```{r slice_cn_by_group}
ukb_br_ci_adj <- bind_rows(
  ukb_covars_br |>
    inner_join(ukb_adj_47s, by = 'eid') |>
    group_by(the_group=NLR_group) |>
    summarise(value = mean(adj_CN),
              SD = sd(adj_CN),
              N = n_distinct(eid),
              margin = qt(ci,df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
    mutate(the_ratio = 'NLR'),
  ukb_covars_br |>
    inner_join(ukb_adj_47s, by = 'eid') |>
    group_by(the_group=PLR_group) |>
    summarise(value = mean(adj_CN),
              SD = sd(adj_CN),
              N = n_distinct(eid),
              margin = qt(ci,df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
    mutate(the_ratio = 'PLR'),
  ukb_covars_br |>
    inner_join(ukb_adj_47s, by = 'eid') |>
    group_by(the_group=SII_group) |>
    summarise(value = mean(adj_CN),
              SD = sd(adj_CN),
              N = n_distinct(eid),
              margin = qt(ci,df=N-1)*SD/sqrt(N),
              value_low = value - margin,
              value_high = value + margin,
              .groups = 'drop') |>
    mutate(the_ratio = 'SII'))
```

```{r plot_cn_by_ratio_group}
blood_ratios_ci_adj_narrow_p <- ukb_br_ci_adj |>
  ggplot(aes(y = the_group, fill = the_group, 
             xmin = value_low, xmax = value_high, x = value)) +
  geom_crossbar(colour = 'grey20', alpha = 0.8, size = 0.75) +
  scale_x_continuous(limits = c(1.55e-4, 1.65e-4),
                     breaks = c(1.55e-4, 1.6e-4, 1.65e-4),
                     labels = fancy_scientific) +
  facet_grid(rows = vars(the_ratio)) +
  scale_fill_brewer(palette = 'Reds') +
  labs(x = bquote(bold("SC-Adjusted 18S Ratio (Mean" %+-% "95% CI)"))) +
  scale_y_discrete('Trait group')  +
  theme_pub() +
  theme(legend.position = 'none')
```

```{r save_cn_by_ratio_group_plot}
ggsave(
  plot = blood_ratios_ci_adj_narrow_p,
  filename = file.path('..', 'figures', '2.G.- Staircase.png'), 
  height = 10, width = 5)
```
