---
title: "Validation"
---

```{r setup}
source('0.0.- Setup.R')
```

## A.- PHESANT comparisons

### CN

```{r read_phesant_cn}
phesant_cn <- read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_with_bc', 'cn', 'results-combined.txt'),
  col_types = cols()) |>
  clean_phesant()
```

### NLR

```{r read_phesant_nlr}
phesant_nlr <- read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_with_bc', 'nlr', 'results-combined.txt'),
  col_types = cols()) |>
  clean_phesant() |>
  filter(! grepl('Blood count', Path))
```

```{r combine_phesant_nlr_cn}
phesant_nlr_cn <- phesant_nlr |>
  inner_join(phesant_cn |>
               select(
                 -varType, -starts_with('Cat'), -resType,
                 -isTraitOfInterest, -description, -Path),
             by = c('varName'), suffix = c('_nlr', '_cn')) |>
  relocate(description) |>
  select(-isTraitOfInterest) |> 
  relocate(starts_with('res'), .after = 22L) |>
  relocate(starts_with('Cat'), .after = 22L) |>
  relocate(Path, .after = 22L)
```

```{r plot_phesant_nlr_cn}
phesant_nlr_cn_p <- phesant_nlr_cn |>
  mutate(description = case_when(
    grepl('F06', description) ~ 'F06: Other mental disorders [...]',
    grepl('D73', description) ~ 'D73: Diseases of spleen',
    TRUE ~ description)) |> 
  mutate(variable = 'NLR',
         fdr_nlr = p.adjust(pvalue_nlr, method = 'fdr'),
         fdr_cn = p.adjust(pvalue_cn, method = 'fdr'),
         the_label = if_else(
           fdr_nlr < 0.01 & fdr_cn < 0.01, 
           description, '')) |>
  ggplot(aes(x = -log10(fdr_nlr),
             y = -log10(fdr_cn))) + 
  ggrepel::geom_text_repel(aes(label = the_label), seed = 500,
                           point.size = 2, size = 5,
                           xlim = c(-log10(0.01), 300),
                           ylim = c(-log10(0.01), 10)) +
  ggrastr::geom_point_rast(
    stroke = 0.75, alpha = 0.5, shape = 21, fill = 'grey20', size = 2) +
  geom_vline(xintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  facet_grid(cols = vars(variable)) +
  scale_x_continuous(limits = c(0, 300)) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = bquote(bold(-log[10](FDR[NLR]))),
       y= bquote(bold(-log[10](FDR[rDNA])))) +
  theme_pub()
```

### PLR

```{r read_phesant_plr}
phesant_plr <- read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined_with_bc', 'plr', 'results-combined.txt'),
  col_types = cols()) |>
  clean_phesant() |>
  filter(! grepl('Blood count', Path))
```

```{r combine_phesant_plr_cn}
phesant_plr_cn <- phesant_plr |>
  inner_join(phesant_cn |>
               select(
                 -varType, -starts_with('Cat'), -resType,
                 -isTraitOfInterest, -description, -Path),
             by = c('varName'), suffix = c('_plr', '_cn')) |>
  relocate(description) |>
  select(-isTraitOfInterest) |> 
  relocate(starts_with('res'), .after = 22L) |>
  relocate(starts_with('Cat'), .after = 22L) |>
  relocate(Path, .after = 22L)
```

```{r plot_phesant_plr_cn}
phesant_plr_cn_p <- phesant_plr_cn |> 
  mutate(variable = 'PLR',
         fdr_plr = p.adjust(pvalue_plr, method = 'fdr'),
         fdr_cn = p.adjust(pvalue_cn, method = 'fdr'),
         the_label = if_else(
           fdr_plr < 0.01 & fdr_cn < 0.01, 
           description, '')) |>
  ggplot(aes(x = -log10(fdr_plr),
             y = -log10(fdr_cn))) + 
  ggrepel::geom_text_repel(aes(label = the_label), seed = 1,
                           point.size = 1, size = 5,
                           xlim = c(-log10(0.01), Inf),
                           ylim = c(-log10(0.01), 9.5)) +
  ggrastr::geom_point_rast(
    stroke = 0.75, alpha = 0.5, shape = 21, fill = 'grey20', size = 2) +
  geom_vline(xintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  facet_grid(cols = vars(variable)) +
  scale_x_continuous(limits = c(0, 300)) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = bquote(bold(-log[10](FDR[PLR]))),
       y= bquote(bold(-log[10](FDR[rDNA])))) +
  theme_pub()
```

### SII

```{r read_phesant_sii}
phesant_sii <- read_tsv(
  file.path(root_fd, 'data', 'phesant',
            'combined', 'sii', 'results-combined.txt'),
  col_types = cols()) |>
  clean_phesant() |>
  filter(! grepl('Blood count', Path))
```

```{r combine_phesant_sii_cn}
phesant_sii_cn <- phesant_sii |>
  inner_join(phesant_cn |>
               select(
                 -varType, -starts_with('Cat'), -resType,
                 -isTraitOfInterest, -description, -Path),
             by = c('varName'), suffix = c('_sii', '_cn')) |>
  relocate(description) |>
  select(-isTraitOfInterest) |> 
  relocate(starts_with('res'), .after = 22L) |>
  relocate(starts_with('Cat'), .after = 22L) |>
  relocate(Path, .after = 22L)
```

```{r plot_phesant_sii_cn}
phesant_sii_cn_p <- phesant_sii_cn |>
  mutate(description = case_when(
    grepl('F06', description) ~ 'F06: Other mental disorders [...]',
    grepl('D73', description) ~ 'D73: Diseases of spleen',
    TRUE ~ description)) |> 
  mutate(variable = 'SII',
         fdr_sii = p.adjust(pvalue_sii, method = 'fdr'),
         fdr_cn = p.adjust(pvalue_cn, method = 'fdr'),
         the_label = if_else(
           fdr_sii < 0.01 & fdr_cn < 0.01, 
           description, '')) |>
  ggplot(aes(x = -log10(fdr_sii),
             y = -log10(fdr_cn))) + 
  ggrepel::geom_text_repel(aes(label = the_label),
                           seed = 1000,
                           size = 5,
                           point.size = 2,
                           xlim = c(-log10(0.01), Inf),
                           ylim = c(-log10(0.01), 9.5)) +
  ggrastr::geom_point_rast(
    stroke = 0.75, alpha = 0.5, size = 2, shape = 21, fill = 'grey20') +
  geom_vline(xintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.01),
             colour = 'grey20',
             alpha = 0.7,
             linetype = 'dashed') +
  facet_grid(cols = vars(variable)) +
  scale_x_continuous(limits = c(0, 300)) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = bquote(bold(-log[10](FDR[SII]))),
       y= bquote(bold(-log[10](FDR[rDNA])))) +
  theme_pub()
```

## Joined Ratios

```{r join_phesant_plots}
phesant_ratios_p <- (
  (phesant_nlr_cn_p +
     theme(plot.margin = margin(r=-200))) | 
  (phesant_plr_cn_p + 
     theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank(),
           axis.title.y = element_blank(),
           plot.margin = margin(r=-200, l = -200))) | 
  (phesant_sii_cn_p + 
     theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank(),
           axis.title.y = element_blank(),
           plot.margin = margin(l=-200)))
)
```

```{r save_phesant_plots}
ggsave(
  plot = phesant_ratios_p,
  filename = file.path('..', 'figures', '3.A.- PHESANT ratios.png'), 
  height = 4, width = 11)
```

```{r overlap_phesant_function}
find_overlaps <- function(th) {
  bind_rows(
  phesant_nlr_cn |>
    mutate(ratio = 'NLR',
           fdr_ratio= p.adjust(pvalue_nlr, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr')),
  phesant_sii_cn |>
    mutate(ratio = 'SII',
           fdr_ratio= p.adjust(pvalue_sii, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr')),
  phesant_plr_cn |>
    mutate(ratio = 'PLR',
           fdr_ratio= p.adjust(pvalue_plr, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr'))) |>
    filter(fdr_cn < th, fdr_ratio < th) |>
    arrange(fdr_cn) |>
    select(description) |>
    distinct()
}
```


```{r overlap_phesant}
ratio_overlaps <- bind_rows(
  phesant_nlr_cn |>
    mutate(ratio = 'NLR',
           fdr_ratio= p.adjust(pvalue_nlr, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr')),
  phesant_sii_cn |>
    mutate(ratio = 'SII',
           fdr_ratio= p.adjust(pvalue_sii, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr')),
  phesant_plr_cn |>
    mutate(ratio = 'PLR',
           fdr_ratio= p.adjust(pvalue_plr, method = 'fdr'),
           fdr_cn = p.adjust(pvalue_cn, method = 'fdr'))) |>
    filter(fdr_cn < 0.1, fdr_ratio < 0.1) |>
  select(ratio, description, fdr_ratio, fdr_cn)
```

## B.- Subpopulations Significance

### Covariates

```{r read_covars}
ukb_gwas_covars <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_gwas_covars.csv'),
  col_types = cols())
```

```{r read_blood_counts}
ukb_blood_counts <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_blood_counts.csv'),
  col_types = cols())
```

```{r construct_blood_count_covars}
ukb_covars_bc <- ukb_gwas_covars |>
  inner_join(ukb_blood_counts, by = 'eid') |>
  select(-Height, -starts_with('Blood'))
```

```{r read_demographics}
ukb_demo <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_demo.csv'),
  col_types = cols())
```

```{r read_blood_ratios}
ukb_br <- read_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_blood_ratios.csv'),
    col_types = cols())
```

```{r construct_blood_ratio_covars}
ukb_covars_br <- ukb_gwas_covars |> 
  inner_join(ukb_br, by = 'eid')
```

### Subpopulations

#### Filtered

```{r read_filtered_subp}
ukb_filtered <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'filt_covars_wb.csv'),
  col_types = cols())
```

```{r construct_filtered_covars}
ukb_bc_filt <- ukb_covars_bc |> filter(
  eid %in% ukb_filtered$eid)
```

#### Unrelated

```{r read_families}
ukb_families <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_families.csv'),
  col_types = cols())
```

```{r identify_unrelated}
ukb_demo_unrel <- bind_rows(
  ukb_demo |>
    inner_join(ukb_families, by = 'eid') |>
    group_by(fam_id) |>
    arrange(fam_id, Birth_Year) |>
    filter(eid == first(eid)) |>
    ungroup(),
  ukb_demo |> filter(
    ! eid %in% ukb_families$eid)) |>
  arrange(eid)
```

```{r construct_unrelated_covars}
ukb_bc_filt_unrel <- ukb_covars_bc |> filter(
  eid %in% ukb_filtered$eid,
  eid %in% ukb_demo_unrel$eid)
```

#### Unmedicated

```{r read_medications}
ukb_meds <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_meds.csv'),
  col_types = cols())
```

```{r read_other_medications}
ukb_meds_o <- ukb_meds |> select(eid, med_other) |>
  filter(med_other == 1)
```

```{r extract_medication_codes}
ukb_meds_code <- ukb_meds |> select(-med_other) |>
  pivot_longer(-eid, values_to = 'medcode') |>
  select(-name) |>
  filter(!is.na(medcode)) |>
  mutate(medcode = as.character(medcode))
```

```{r extract_med_hits}
phesant_meds <- phesant_nlr_cn |>
  arrange(pvalue_cn) |>
  select(description, varName, pvalue_cn) |>
  filter(pvalue_cn < 1e-2, grepl("^20003#", varName)) |>
    mutate(medcode = gsub("20003#", '', varName))
```

```{r find_medicated}
ukb_medicated <- unique(c(
  ukb_meds_code |> 
    inner_join(phesant_meds, by = 'medcode') |>
    pull(eid),
  ukb_meds_o |> pull(eid)))
```


```{r construct_unmedicated_covars}
ukb_bc_filt_unmed <- ukb_covars_bc |> filter(
  eid %in% ukb_filtered$eid,
  ! eid %in% ukb_medicated)
```

#### Healthy

```{r read_healthy}
ukb_healthy <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'healthy_covars_wb.csv'),
  col_types = cols())
```

```{r construct_healthy_covars}
ukb_bc_healthy <- ukb_covars_bc |> filter(
  eid %in% ukb_healthy$eid)
```

#### Healthy Unmedicated

```{r construct_healthy_unmedicated_covars}
ukb_bc_healthy_unmed <- ukb_covars_bc |> filter(
  eid %in% ukb_healthy$eid,
  ! eid %in% ukb_medicated)
```

#### Subpopulation titles

```{r create_subpopulation_names}
subpopulations <- c(
  'F' = paste0(
    '(i) Filtered\n \n',
    '[N = ', format(nrow(ukb_bc_filt), big.mark = ","), ']'),
  'FU'=paste0(
    '(ii) Filtered\nUnrelated\n',
    '[N = ', format(nrow(ukb_bc_filt_unrel), big.mark = ","), ']'),
  'FM'=paste0(
    '(iii) Filtered\nUnmedicated\n',
    '[N = ', format(nrow(ukb_bc_filt_unmed), big.mark = ","), ']'),
  'H'=paste0(
    '(iv) Healthy\n \n',
    '[N = ', format(nrow(ukb_bc_healthy), big.mark = ","), ']'),
  'HM'=paste0(
    '(v) Healthy\nUnmedicated\n',
    '[N = ', format(nrow(ukb_bc_healthy_unmed), big.mark = ","), ']'))
```

### Associations

#### Functions

```{r fit_blood_count_function}
fit_bc_scale <- function(df, subp = 'F', variable) {
  flm <- lm(as.formula(paste0('`', variable, '`',  '~ .')) , df |>
              mutate(across(where(is.numeric), scale)) |>
              select(Sex, starts_with('Age'), contains('Centre'), 
                     starts_with('PC'), starts_with('Adj'),
                     all_of(variable),
                     CN))
  broom::tidy(flm, conf.int = TRUE) |> 
    mutate(
      pop = subpopulations[subp],
      lpval = -log10(p.value),
      scoeff = sign(estimate)) |>
    filter(term == 'CN') |>
    mutate(term = variable) |>
    relocate(term)
}
```

```{r fit_blood_counts_function}
fit_bcs <- function(df, subp = 'F') {
  all_vars <- colnames(df)
  variables <- all_vars[grepl('#', all_vars)]
  bind_rows(lapply( X = variables, 
                    FUN = function(x) fit_bc_scale(df, subp, x) ) )
}
```

```{r fit_blood_ratios_subp_function}
fit_subp_ratio <- function(df, subp = 'F') {
  lapply(X = c('NLR', 'PLR', 'LMR', 'SII'),
         FUN = function(x) fit_ratio_scale(df, x) |>
           mutate(pop = subpopulations[subp]))
}
```

```{r fit_blood_count_biochem_function}
fit_bc_biochem_scale <- function(df, subp = 'F', variable) {
  flm <- lm(as.formula(paste0('`', variable, '`',  '~ .')) , df |>
              mutate(across(where(is.numeric), scale)) |>
              select(Sex, starts_with('Age'), contains('Centre'), 
                     starts_with('PC'), starts_with('Adj'),
                     all_of(variable), all_of(ukb_biochem_vars),
                     CN))
  broom::tidy(flm, conf.int = TRUE) |> 
    mutate(
      pop = subpopulations[subp],
      lpval = -log10(p.value),
      scoeff = sign(estimate)) |>
    filter(term == 'CN') |>
    mutate(term = variable) |>
    relocate(term)
}
```

```{r fit_blood_counts_biochem_function}
fit_bcs_biochem <- function(df, subp = 'F') {
  all_vars <- colnames(df)
  variables <- all_vars[grepl('#', all_vars)]
  bind_rows(lapply( X = variables, 
                    FUN = function(x) fit_bc_biochem_scale(df, subp, x) ) )
}
```

```{r fit_blood_ratio_biochem_function}
fit_ratio_biochem <- function(df, the_ratio = 'NLR') {
  ratios_remove <- setdiff(
    c('NLR', 'PLR', 'LMR', 'SII'), the_ratio)
  broom::tidy(lm.beta::lm.beta(lm(CN ~ . , df |>
                mutate(across(where(is.numeric), scale)) |>
               select(-eid, -any_of(ratios_remove)))),
              conf.int = TRUE) |>
  mutate(lpval = -log10(p.value),
         scoeff = sign(std_estimate)) |>
  filter(term == the_ratio)
}
```

```{r fit_blood_ratios_subp_biochem_function}
fit_ratios_biochem <- function(df, subp = 'F') {
  lapply(X = c('NLR', 'PLR', 'LMR', 'SII'),
         FUN = function(x) fit_ratio_biochem(df, x) |>
           mutate(pop = subpopulations[subp]))
}
```

#### No Biochemistry Covariates

```{r fit_blood_counts_no_biochemistry}
ukb_subp_bc <- bind_rows(
  fit_bcs(ukb_covars_bc |>
           filter(eid %in% ukb_bc_filt$eid),
         subp = 'F'),
  fit_bcs(ukb_covars_bc |>
           filter(eid %in% ukb_bc_filt_unmed$eid),
         subp = 'FM'),
  fit_bcs(ukb_covars_bc |>
           filter(eid %in% ukb_bc_filt_unrel$eid),
         subp = 'FU'),
  fit_bcs(ukb_covars_bc |>
           filter(eid %in% ukb_bc_healthy$eid),
         subp = 'H'),
  fit_bcs(ukb_covars_bc |>
           filter(eid %in% ukb_bc_healthy_unmed$eid),
         subp = 'HM')) |>
  mutate(type = 'Counts')
```


```{r bit_blood_ratios_no_biochemistry}
ukb_subp_br <- bind_rows(
  fit_subp_ratio(ukb_covars_br |>
                   filter(eid %in% ukb_bc_filt$eid), 'F'), 
  fit_subp_ratio(ukb_covars_br |>
                   filter(eid %in% ukb_bc_filt_unrel$eid), 'FU'),
  fit_subp_ratio(ukb_covars_br |>
                   filter(eid %in% ukb_bc_filt_unmed$eid), 'FM'),
  fit_subp_ratio(ukb_covars_br |>
                   filter(eid %in% ukb_bc_healthy$eid), 'H'),
  fit_subp_ratio(ukb_covars_br |>
                   filter(eid %in% ukb_bc_healthy_unmed$eid), 'HM'))  |> 
  mutate(type = 'Ratios')
```

#### Biochemistry as Covariates

```{r read_selected_biomarkers}
ukb_biochem <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_biochem.csv'),
  col_types = cols())
```

```{r extract_selected_biomarker_names}
ukb_biochem_vars <- colnames(ukb_biochem)[-1]
```

```{r construct_blood_counts_biomarker_cov}
ukb_covars_biochem_bc <- ukb_gwas_covars |>
  inner_join(ukb_blood_counts, by = 'eid') |>
  inner_join(ukb_biochem, by = 'eid') |>
  select(-Height, -starts_with('Blood'))
```

```{r fit_blood_counts_biomarker}
ukb_subp_biochem_bc <- bind_rows(
  fit_bcs_biochem(ukb_covars_biochem_bc |>
           filter(eid %in% ukb_bc_filt$eid),
         subp = 'F'),
  fit_bcs_biochem(ukb_covars_biochem_bc |>
           filter(eid %in% ukb_bc_filt_unmed$eid),
         subp = 'FM'),
  fit_bcs_biochem(ukb_covars_biochem_bc |>
           filter(eid %in% ukb_bc_filt_unrel$eid),
         subp = 'FU'),
  fit_bcs_biochem(ukb_covars_biochem_bc |>
           filter(eid %in% ukb_bc_healthy$eid),
         subp = 'H'),
  fit_bcs_biochem(ukb_covars_biochem_bc |>
           filter(eid %in% ukb_bc_healthy_unmed$eid),
         subp = 'HM')) |>
  mutate(type = 'Counts')
```

```{r construct_blood_ratios_biomarker_cov}
ukb_covars_biochem_br <- ukb_gwas_covars |>
  inner_join(ukb_br, by = 'eid') |>
  inner_join(ukb_biochem, by = 'eid') |>
  select(-Height, -starts_with('Blood'))
```

```{r fit_blood_ratios_biomarker}
ukb_subp_biochem_br <- bind_rows(
  fit_ratios_biochem(ukb_covars_biochem_br |>
                   filter(eid %in% ukb_bc_filt$eid), 'F'), 
  fit_ratios_biochem(ukb_covars_biochem_br |>
                   filter(eid %in% ukb_bc_filt_unrel$eid), 'FU'),
  fit_ratios_biochem(ukb_covars_biochem_br |>
                   filter(eid %in% ukb_bc_filt_unmed$eid), 'FM'),
  fit_ratios_biochem(ukb_covars_biochem_br |>
                   filter(eid %in% ukb_bc_healthy$eid), 'H'),
  fit_ratios_biochem(ukb_covars_biochem_br |>
                   filter(eid %in% ukb_bc_healthy_unmed$eid), 'HM'))  |> 
  mutate(type = 'Ratios')
```

#### Joint Plot

```{r plot_subp_pvalues}
subp_biochem_sig_p <- bind_rows(
  bind_rows(
    ukb_subp_bc,
    ukb_subp_br) |> 
    mutate(Control = 'No'),
  bind_rows(
    ukb_subp_biochem_bc,
    ukb_subp_biochem_br) |> 
    mutate(Control = 'Yes')) |>
  mutate(Control = forcats::fct_rev(Control),
         base_pop = if_else(grepl('Filt', pop),
                            'F', 'H')) |>
  ggplot(aes(y = term, x = lpval, fill = base_pop, alpha = Control)) +
  geom_col(colour = 'grey20', linewidth = 0.75,
           position = position_dodge2()) +
  geom_vline(xintercept = -log10(0.05),
             colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(type), 
             cols = vars(pop),
             scales = 'free', space = 'free_y') +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_alpha_manual('Control for PHESANT Blood Biochemistry Hits', 
                    values = c('No' = 0.9,
                               'Yes' = 0.3)) +
  scale_fill_manual(values = c('F' = 'burlywood3',
                               'H' = 'slategray3')) +
  theme_pub() +
  theme(legend.position = 'bottom',
        legend.title = element_text(size = 14, face = 'bold'),
        legend.text = element_text(size = 12),
        legend.margin = margin(t = -10),
        axis.text.y = element_text(size = 14),
        strip.text.y = element_blank(),
        panel.spacing = unit(0.1, 'lines')) +
  guides(fill = 'none')
```

```{r save_subp_pvalues_plot}
ggsave(
  plot = subp_biochem_sig_p,
  filename = file.path('..', 'figures', '3.B.- Subpopulations p-values.png'), 
  height = 8, width = 10)
```

## C.- Subpopulations Betas

```{r plot_subp_betas}
subp_betas_p <- bind_rows(
  ukb_subp_bc,
  ukb_subp_br) |> 
  filter(grepl('\\(i\\)|\\(iv\\)', pop)) |>
  mutate(subp = factor(
    if_else(grepl('Filt', pop), 'Filtered', 'Healthy'))) |>
  ggplot(aes(y = term, x = estimate, xmin = conf.low, xmax = conf.high,
             shape = subp)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_pointrange(aes(fill = subp),
                  colour = 'grey30', size = 1,
                linewidth = 0.75, alpha = 0.8,
                position = position_dodge2(width = 0.75)) +
  scale_fill_manual('Subpopulation', 
                    values = c('Filtered' = 'burlywood3',
                               'Healthy' = 'slategray3')) +
  scale_shape_manual('Subpopulation', 
                    values = c('Filtered' = 21,
                               'Healthy' = 23)) +
  scale_x_continuous(limits = c(-0.065, 0.065),
                     breaks = c(-0.05, 0, 0.05)) +
  facet_grid(rows = vars(type), scales = 'free_y', space = 'free') +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  theme_pub() +
  theme(legend.position = 'bottom',
        legend.title = element_text(size = 14, face = 'bold'),
        legend.text = element_text(size = 13),
        legend.margin = margin(t = -10),
        strip.text.y = element_blank(),
        axis.text.y = element_text(size = 14),
        panel.spacing = unit(0.1, 'lines'),
        plot.margin = margin(r = 10))
```

```{r save_subp_betas_plot}
ggsave(
  plot = subp_betas_p,
  filename = file.path('..', 'figures', '3.C.- Subpopulations betas.png'), 
  height = 7, width = 5.2)
```

## D.- Other Ethnicities, combined

```{r read_all_cn}
ukb_47s <- read_csv(
  file.path(root_fd, 'data', 'cn', 'ukb_47s.csv'),
  col_types = cols())
```

```{r extract_non_wb_cn}
ukb_47s_o <- ukb_demo %>%
  inner_join(ukb_47s, by = 'eid') %>%
  filter(Eth_Back != 1001)
```

```{r read_genotyping_array}
ukb_geno_array <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_array.csv'),
  col_types = cols()
)
```

```{r read_pcs}
ukb_pcs <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_pc.csv'),
  col_types = cols()
)
```

```{r read_seq_rel}
ukb_sr <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_sr.csv'),
  col_types = cols()
)
```

```{r read_adj_tel_length}
ukb_atl <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_adj_tl.csv'),
  col_types = cols()
)
```

```{r read_assessment_centre}
ukb_ac <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_ac.csv'),
  col_types = cols()
)
```

```{r construct_all_ratios}
ukb_br_all <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_blood_counts.csv'),
  col_types = cols()) |>
  filter(`Monocyte #` > 0, !is.na(`Platelet #`)) |>
  mutate(NLR = `Neutrophil #` / `Lymphocyte #`,
         PLR = `Platelet #` / `Lymphocyte #`,
         LMR = `Lymphocyte #` / `Monocyte #`,
         SII = (`Neutrophil #` *`Platelet #`) / `Lymphocyte #`)
```

```{r construct_non_wb_covars}
ukb_covars_o <- ukb_47s_o %>%
    filter(eid %in% c(ukb_geno_array %>% 
                        filter(Array == 1) %>% pull(eid))) %>%
    inner_join(ukb_pcs, by = 'eid') %>%
    inner_join(ukb_sr, by = 'eid') %>%
    inner_join(ukb_atl, by = 'eid') %>%
    inner_join(ukb_ac %>% select(-coding), by = 'eid') %>%
    inner_join(ukb_br_all, by = 'eid') %>%
    mutate(`Age Sq` = Age * Age) %>%
    select(eid, Sex, 
           starts_with('Age'), 
           Eth_Back,
           contains('Centre'),
           paste0('PC', 1:40),
           starts_with('Adj'),
           ends_with('#'),
           NLR, PLR, SII, LMR,
           CN)

ukb_covars_o[is.na(ukb_covars_o) | ukb_covars_o == "Inf"] <- NA
```

```{r fit_non_wb_blood_function}
fit_o <- function(df, the_variable = 'Neutrophil #',
                  num_pc = 10) {
  broom::tidy(
    lm(as.formula(paste0('`', the_variable, '`',  '~ .')) , 
       df |>
         mutate(across(where(is.numeric), scale)) |>
         select(Sex, starts_with('Age'), 
                Eth_Back, contains('Centre'), 
                any_of(paste0('PC', 1:num_pc)),
                starts_with('Adj'),
                all_of(the_variable),
                CN)),
    conf.int = TRUE) |> 
    mutate(lpval = -log10(p.value),
           scoeff = sign(estimate),
           term = gsub('`', '', term)) |>
    filter(term == 'CN') |>
    mutate(term = the_variable) |>
    relocate(term)
}
```


```{r fit_non_wb_blood}
bc_vars <- colnames(ukb_covars_o)[grepl('#', colnames(ukb_covars_o))]
other_eth_sig <- bind_rows(
  bind_rows(
  lapply(X = bc_vars,
       FUN = function(x) fit_o(ukb_covars_o, x))) |>
    mutate(type = 'Counts'),
  bind_rows(
  lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_o(ukb_covars_o, x))) |>
    mutate(type = 'Ratios') )
```

```{r plot_non_wb_betas}
other_eth_betas_p <- other_eth_sig |> 
  ggplot(aes(y = term, x = estimate, xmin = conf.low,
             xmax = conf.high, fill = as.factor(scoeff))) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_pointrange(colour = 'grey30',
                linewidth = 0.75, alpha = 0.8, shape = 23) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  facet_grid(rows = vars(type), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(-0.06, 0.06),
                     breaks = c(-0.05, 0, 0.05)) +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  theme_pub() +
  theme(legend.position = 'none',
        strip.text.y = element_blank(),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13)) 
```

```{r plot_non_wb_pvalues}
other_eth_sig_p <- other_eth_sig |> 
  ggplot(aes(y = term, x = lpval, fill = as.factor(scoeff))) +
  geom_col(alpha = 0.8, colour = 'grey20', linewidth = 0.75) +
  geom_vline(xintercept = -log10(0.05), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(type), 
             scales = 'free_y', space = 'free') +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_fill_manual('', 
                    values = c('-1' = 'lightcoral',
                               '1' = 'royalblue4')) +
  theme_pub() +
  theme(legend.position = 'none',
        strip.text.y = element_blank(),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13)) 
```

```{r save_non_wb_associations}
ggsave(
  plot = (other_eth_betas_p | (
    other_eth_sig_p +
      theme(axis.text.y = element_blank(),
            axis.title.y = element_blank(),
            axis.ticks.y = element_blank()))
    ) + plot_annotation(
      title = 'All Ethnicities other than White British [N = 22,609]',
      theme = theme(plot.title = element_text(hjust = 0.55))) +
  plot_layout(widths = c(0.6, 0.4)),
  filename = file.path('..', 'figures', '3.D.- Other Ethnicities.png'), 
  height = 5, width = 6)
```

```{r define_format_comma}
my_comma <- scales::label_comma(accuracy = 1)
```

## E.- Individual Countries and Ethnicities

```{r fit_uk_country}
ukb_countries <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_countries.csv'), 
            col_types = cols())
ukb_gwas_country <- ukb_br_all |>
  inner_join(ukb_gwas_covars, by = 'eid') |>
  inner_join(ukb_countries, by = 'eid')

wb_nation_sig <- bind_rows(
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_gwas_country |>
                                     filter(country == 'England'), x))) |>
    mutate(
      eth = paste0(
        'White British in England\n[N = ', 
        my_comma(
          nrow(ukb_gwas_country |> filter(country == 'England'))), ']'),
      eth_order = 1),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_gwas_country |>
                                     filter(country == 'Scotland'), x))) |>
    mutate(
      eth = paste0(
        'White British in Scotland\n[N = ', 
        my_comma(
          nrow(ukb_gwas_country |> filter(country == 'Scotland'))), ']'),
      eth_order = 2),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_gwas_country |>
                                     filter(country == 'Wales'), x))) |>
    mutate(
      eth = paste0(
        'White British in Wales\n[N = ', 
        my_comma(
          nrow(ukb_gwas_country |> filter(country == 'Wales'))), ']'),
      eth_order = 3))
```

```{r fit_sel_eth}
specific_eth_sig <- bind_rows(
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back == 5), x))) |>
    mutate(
      eth = paste0(
        'Chinese\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back == 5))), ']'),
      eth_order = 9),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back == 1002), x))) |>
    mutate(
      eth = paste0(
        'White Irish\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back == 1002))), ']'),
      eth_order = 4),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back == 1003), x))) |>
    mutate(
      eth = paste0(
        'White Other\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back == 1003))), ']'),
      eth_order = 5),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back > 3000,
                                             Eth_Back < 4000), x))) |>
    mutate(
      eth = paste0(
        'Asian or Asian British\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back > 3000, Eth_Back < 4000))),
        ']'),
      eth_order = 7),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back > 4000,
                                             Eth_Back < 5000), x))) |>
    mutate(
      eth = paste0(
        'Black or Black British\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back > 4000, Eth_Back < 5000))),
        ']'),
      eth_order = 6),
  bind_rows(
    lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_ratio_scale(ukb_covars_o |> 
                                      filter(Eth_Back > 2000,
                                             Eth_Back < 3000), x))) |>
    mutate(
      eth = paste0(
        'Mixed\n[N = ', 
        my_comma(
          nrow(ukb_covars_o |> filter(Eth_Back > 2000, Eth_Back < 3000))),
        ']'),
      eth_order = 10)) |> 
  mutate(term = forcats::fct_rev(term))
```

```{r plot_sel_eth_betas}
all_eth_beta_p <- bind_rows(
  wb_nation_sig,
  specific_eth_sig) |>
  filter(term %in% c('SII', 'NLR')) |>
  arrange(eth_order) |>
  mutate(eth = forcats::fct_rev(forcats::fct_inorder(eth))) |>
  ggplot(aes(y = eth, x = estimate, xmin = conf.low, xmax = conf.high,
             group = term)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_pointrange(
    aes(shape = term, fill = term), 
    colour = 'grey30', linewidth = 0.75, alpha = 0.8,
    position = position_dodge2(width = 0.65)) +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  scale_x_continuous(limits = c(-0.05, 0.25),
                     breaks = c(0, 0.1, 0.2)) +
  scale_shape_manual('', values = c('NLR' = 21, 'SII' = 23)) +
  scale_fill_manual('',
                    values = c('NLR' = 'steelblue1', 'SII' = 'royalblue4')) +
  theme_pub() +
  theme(legend.position = 'bottom',
        strip.text = element_text(size = 14),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13))
```

```{r plot_sel_eth_pvalues}
all_eth_sig_p <- bind_rows(
  wb_nation_sig,
  specific_eth_sig) |>
  filter(term %in% c('SII', 'NLR')) |>
  arrange(eth_order) |>
  mutate(eth = forcats::fct_rev(forcats::fct_inorder(eth))) |>
  ggplot(aes(y = eth, x = lpval, group = term)) +
  geom_vline(xintercept = -log10(0.05), linewidth = 0.5, colour = 'grey80') +
  geom_col(aes(fill = term), 
    colour = 'grey30', linewidth = 0.75, alpha = 0.8,
    position = position_dodge2(width = 0.65)) +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_fill_manual('',
                    values = c('NLR' = 'steelblue1', 'SII' = 'royalblue4')) +
  theme_pub() +
  theme(legend.position = 'bottom',
        strip.text = element_text(size = 14),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13))
```

```{r save_sel_eth_plot}
ggsave(
  plot = (all_eth_beta_p | (
    all_eth_sig_p +
      theme(axis.text.y = element_blank(),
            axis.title.y = element_blank(),
            axis.ticks.y = element_blank()))
    ) +
  plot_layout(widths = c(0.6, 0.4)),
  filename = file.path('..', 'figures', '3.E.- Specific Ethnicities.png'), 
  height = 5.5, width = 8)
```

## S17.- Number of Principal Components

```{r fit_40_pc}
other_eth_sig_40_pc <- bind_rows(
  bind_rows(
  lapply(X = bc_vars,
       FUN = function(x) fit_o(ukb_covars_o, x, num_pc = 40))) |>
    mutate(type = 'Counts'),
  bind_rows(
  lapply(X = c('NLR', 'PLR', 'SII', 'LMR'),
       FUN = function(x) fit_o(ukb_covars_o, x, num_pc = 40))) |>
    mutate(type = 'Ratios') )
```

```{r plot_40_pc_betas}
other_eth_betas_40_pc_p <- bind_rows(
  other_eth_sig_40_pc |> mutate(num_pc = '40 PC'),
  other_eth_sig |> mutate(num_pc = '10 PC')) |> 
  ggplot(aes(y = term, x = estimate, xmin = conf.low,
             xmax = conf.high, fill = num_pc, group = num_pc)) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = 'grey80') +
  geom_pointrange(colour = 'grey30',
                linewidth = 0.75, alpha = 0.8, shape = 23,
                position = position_dodge(width = 0.9)) +
  scale_fill_manual('Number of\nPrincipal Components', 
                    values = c('10 PC' = 'grey50',
                               '40 PC' = 'grey95')) +
  facet_grid(rows = vars(type), scales = 'free_y', space = 'free') +
  scale_x_continuous(limits = c(-0.06, 0.06),
                     breaks = c(-0.05, 0, 0.05)) +
  labs(y = '', x = "Standardised Beta, 95% CI") +
  theme_pub() +
  theme(legend.position = 'none',
        strip.text.y = element_blank(),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13)) 
```

```{r plot_40_pcs_pvalues}
other_eth_sig_40_pc_p <- bind_rows(
  other_eth_sig_40_pc |> mutate(num_pc = '40 PC'),
  other_eth_sig |> mutate(num_pc = '10 PC')) |>  
  ggplot(aes(y = term, x = lpval, fill = num_pc, group = num_pc)) +
  geom_col(alpha = 0.8, colour = 'grey20', linewidth = 0.75,
            position = position_dodge(width = 0.9)) +
  geom_vline(xintercept = -log10(0.05), colour = 'grey60', alpha = 0.8,
             linewidth = 0.75, linetype = 'dashed') +
  facet_grid(rows = vars(type), 
             scales = 'free_y', space = 'free') +
  labs(y = '', x = bquote(bold(-log[10](p-value)))) +
  scale_fill_manual('Number of\nPrincipal\nComponents', 
                    values = c('10 PC' = 'grey50',
                               '40 PC' = 'grey95')) +
  theme_pub() +
  theme(
        strip.text.y = element_blank(),
        panel.spacing = unit(0.1, 'lines'),
        axis.title = element_text(size = 13)) 
```

```{r save_40_pcs_plot}
ggsave(
  plot = (other_eth_betas_40_pc_p | (
    other_eth_sig_40_pc_p +
      theme(axis.text.y = element_blank(),
            axis.title.y = element_blank(),
            axis.ticks.y = element_blank()))
    ) + plot_annotation(
      title = 'All Ethnicities other than White British [N = 22,609]',
      theme = theme(plot.title = element_text(hjust = 0.55),
                    plot.subtitle = element_text(hjust = 0.55))),
  filename = file.path('..', 'figures_supp', 
                       'S17.- Other Ethnicities, 40 PCs.png'), 
  height = 6, width = 8)
```
