---
title: "0.1.- Generate phenotype tables from UK Biobank releases"
---

```{r setup}
source('0.0.- Setup.R')
```

## Initial Releases

```{r read_last_release}
ukb_pheno_last <- read_csv(
  file.path(root_fd, 'data', 'releases', 'pheno_9.csv'),
  col_types = cols())
```

```{r read_pheno, warning = F}
ukb_pheno_initial <- 
  read_csv(file.path(root_fd, 'data', 'metadata', 'ukb51456.csv'),
           col_types=cols()) |>
  inner_join(
    read_csv(file.path(root_fd, 'data', 'metadata', 'ukb51638.csv'),
           col_types=cols()),
    by = 'eid') |>
  inner_join(
    read_csv(file.path(root_fd, 'data', 'metadata', 'ukb51937.csv'),
           col_types=cols()),
    by = 'eid') |>
  inner_join(
    read_csv(file.path(root_fd, 'data', 'metadata', 'ukb668859.csv'),
           col_types=cols()),
    by = 'eid') |>
  inner_join(
    read_csv(file.path(root_fd, 'data', 'metadata', 'ukb669081.csv'),
             col_types=cols()),
    by='eid') |>
  filter(eid %in% ukb_pheno_last$eid)
```

## Assessment Centre

```{r}
ukb_ac <- ukb_pheno_initial |>
  select(eid, `coding`=`54-0.0`) |>
  inner_join(
    read_tsv(file.path(root_fd, 'data', 'metadata', 'ukb_coding_ac.tsv'),
             col_types = cols()),
    by = c('coding')) |>
  rename(`As Centre` = meaning)
```

```{r}
ukb_ac |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_ac.csv'),
  na = '')
```

## Demographic Variables

```{r demo}
ukb_demo <- ukb_pheno_initial |> select(
  eid,
  Sex = starts_with('31-'),
  Age = starts_with('21003-'),
  Birth_Year=starts_with('34-'),
  Eth_Back = starts_with('21000-'), 
  Eth_Back_Gen = starts_with('22006-')) |>
  mutate(Sex = if_else(Sex == 0, 'Female', 'Male')) |>
  select(eid, Sex, Age = Age1, Birth_Year,
         Eth_Back_Gen, Eth_Back = Eth_Back1)
```

```{r}
ukb_demo |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_demo.csv'),
  na = '')
```

```{r}
ukb_countries <- ukb_pheno_initial |> select(
  eid,
  non_uk=starts_with('20115-'),
  uk=starts_with('1647-0')) |>
  left_join(
    read_tsv(
      file.path(root_fd, 'data', 'pheno', 'ukb_non_uk_country_coding.tsv'),
      col_types = cols()),
    by = c('non_uk'='coding')) |>
  left_join(
    read_tsv(
      file.path(root_fd, 'data', 'pheno', 'ukb_uk_country_coding.tsv'),
      col_types = cols()),
    by = c('uk'='coding')) |>
  left_join(
    read_tsv(
      file.path(root_fd, 'data', 'pheno', 'ukb_non_uk_country_coding.tsv'),
      col_types = cols()) |>
      select(coding, region = meaning),
    by = c('parent_id'='coding')) |>
  mutate(country = if_else(meaning.y == 'Elsewhere', meaning.x, meaning.y),
         region = if_else(meaning.y == 'Elsewhere', region, 'UK')) |>
  select(eid, region, country)
```

```{r}
ukb_countries |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_countries.csv'),
  na = '')
```

## Biometric Variables

```{r}
ukb_bm <- ukb_pheno_initial |>
  select(
    eid,
    Height = `50-0.0`,
    BMI = `21001-0.0`)
```

```{r}
ukb_bm |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_body.csv'),
  na = '')
```

#### Adjusted Telomere Lengths

```{r}
ukb_atl <- ukb_pheno_initial |> 
  select(eid, `Adj TL`=starts_with('22191-0.0'))
```

```{r}
ukb_atl |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_adj_tl.csv'),
  na = '')
```

## Principal Components

```{r pcs}
ukb_pcs <- ukb_pheno_initial |>
  select(
    eid,
    PC = starts_with('22009-'))
```

```{r}
ukb_pcs |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_pc.csv'),
  na = '')
```

## Blood Composition

### Counts and Percentages

```{r blood_perc}
ukb_pheno_blood_perc <- read_csv(
  file.path(root_fd, 'data' ,'metadata', 'ukb_blood_cell_perc.csv'),
  col_types=cols()) |>
  select(eid, ends_with('-0.0')) |>
  filter(eid %in% ukb_pheno_last$eid)
```

```{r}
ukb_pheno_blood <- ukb_pheno_initial |> 
  inner_join(ukb_pheno_blood_perc, by = 'eid') |>
  select(
    eid,
    `Platelet #` = `30080-0.0`,
    `Erythrocyte #` = `30010-0.0`,
    `Lymphocyte #` = `30120-0.0`,
    `Lymphocyte %` = `30180-0.0`,
    `Monocyte #` = `30130-0.0`,
    `Monocyte %` = `30190-0.0`,
    `Neutrophil #` = `30140-0.0`,
    `Neutrophil %` = `30200-0.0`,
    `Eosinophil #` = `30150-0.0`,
    `Eosinophil %` = `30210-0.0`,
    `Basophil #` = `30160-0.0`,
    `Basophil %` = `30220-0.0`,
    `NRB %` = `30230-0.0`)
```  
  
```{r}
ukb_pheno_blood |>
  select(eid, ends_with('#')) |>
  write_csv(file.path(root_fd, 'data', 'pheno', 'ukb_blood_counts.csv'),
            na = '')
```

### Processing Time

```{r}
ukb_blood_extr <- read_csv(
  file.path(root_fd, 'data', 'releases', 'pheno_9.csv'),
         col_types = cols()) |>
  select(eid, Extraction = `3166-0.0`) |>
  filter(!is.na(Extraction)) |>
  mutate(year = lubridate::year(Extraction)) |>
  filter(as.numeric(as.character(year)) > 1900) |>
  select(-year)
```

```{r}
ukb_blood_at <- read_csv(
  file.path(root_fd, 'data', 'releases', 'ukb_bc_proc_raw.csv'),
         col_types = cols()) |>
  select(eid, `Acquisition Time` = `30142-0.0`)
```

```{r}
ukb_blood_extr |>
  inner_join(ukb_blood_at, by = 'eid') |>
  mutate(Delay = `Acquisition Time` - Extraction) |>
  arrange(desc(Delay)) |>
  filter(as.numeric(Delay) < 36) |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_delay.csv'),
    na = '')
```

### Principal Components

```{r}
ukb_pheno_blood_perc_f <- ukb_pheno_blood |>
  filter(`NRB %` == 0) |>
  select(eid, ends_with('%'), -starts_with('NRB'))
```

```{r}
ukb_pheno_blood_pcs <- FactoMineR::PCA(
  ukb_pheno_blood_perc_f |> select(-eid),
  scale.unit = T, graph = F, ncp = 10)
```

```{r}
save(ukb_pheno_blood_pcs, 
     file = file.path(root_fd, 'data', 'pheno', 'ukb_blood_pcs.Rdata'))
```

```{r}
ukb_blood_pc <- as_tibble(ukb_pheno_blood_pcs$ind$coord) |>
  mutate(eid = ukb_pheno_blood_perc_f$eid) |>
  relocate(eid) |>
  rename_with(~ gsub("Dim.", "Blood PC", .x, fixed = TRUE))
```

```{r}
ukb_blood_pc |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_blood_pc.csv'),
  na = '')
```

### Full list of phenotypes from Vuckovic et al

```{r}
vuckovic_phenotypes <- readxl::read_excel(
  file.path(root_fd, 'data', 'vuckovic', 'blood_pheno.xlsx'),
  skip = 2) |>
  select(Abbreviation = `Standard Abbreviation`,
         Name = `Long Name`,
         Variable = `UKB Variable`) |>
  arrange(Variable)
```

```{r}
vuckovic_variables <- paste0(
  vuckovic_phenotypes$Variable, '-0.0'
)
```

```{r}
ukb_vuckovic <- ukb_pheno_initial |>
  inner_join(ukb_pheno_blood_perc, by = 'eid') |>
  select(eid, any_of(vuckovic_variables)) 

colnames(ukb_vuckovic) <- c('eid', vuckovic_phenotypes$Abbreviation)
```

```{r}
ukb_vuckovic |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_vuckovic.csv'),
    na = '')
```

## Genotyping Array  
  
```{r}
ukb_geno_batch <- ukb_pheno_initial |>
  select(eid, Batch = starts_with('22000-'))
```

```{r}
ukb_geno_array <- ukb_geno_batch |>
  mutate(Array=if_else(Batch > 0, '1', '2'),
         Batch=paste0(Array, '_', abs(Batch))) |>
  select(eid, Array, Batch)
```

```{r}
ukb_geno_array |> 
  write_csv(file.path(root_fd, 'data', 'pheno', 'ukb_array.csv'),
            na = '')
```

```{r}
ukb_geno_array_wb <- ukb_geno_array |>
  inner_join(ukb_demo, by = 'eid') |>  
  filter(Eth_Back_Gen == 1, Eth_Back == 1001) |>
  select(eid, Array, Batch)
```

## Sequencing Centre

```{r sc}
ukb_sr <- bind_rows(
  lapply(
    X = list.files(file.path(root_fd, 'data', 'rg')),
    FUN = function(x)
      read_csv(file.path(root_fd, 'data', 'rg', x), 
               col_types = cols()))) |> 
  rowwise() |>
  mutate(SC = if_else(grepl('CN:SC', rg),
                      'Sanger', 'deCODE'),
         SR = if_else(
           grepl('Main Phase', rg) | SC != 'Sanger',
           SC, paste0(SC, 'Vanguard'))) |>
  separate(sample, into = 'eid',
           sep = '_', extra = 'drop') |>
  mutate(eid = as.numeric(eid), 
         SR = gsub('V', ' V', SR)) |>
  select(eid, `Seq Centre`=SC, `Seq Release`=SR) |>
  filter(eid %in% ukb_pheno_last$eid)
```

```{r}
ukb_sr |> write_csv('../data/pheno/ukb_sr.csv', na = '')
```

## 47S CN

```{r read_47s}
ukb_47s <- read_csv(file.path(root_fd, 'data', 'cn', 'ukb_ratios.csv'),
                    col_types = cols()) |>
  filter(rdna > 0) |>
  rename(CN = rdna_ratio) |>
  filter(eid %in% ukb_pheno_last$eid)
```

```{r}
ukb_47s |>
  write_csv(file.path(root_fd, 'data', 'cn', 'ukb_47s.csv'),
            na = '')
```


```{r 47s_wb}
ukb_47s_wb <- ukb_47s |>
  inner_join(ukb_demo |>
               select(eid, Eth_Back_Gen, Eth_Back),
             by = 'eid') |>
  filter(Eth_Back_Gen == 1, Eth_Back == 1001) |>
  select(-starts_with('Eth_Back'))
```

```{r}
ukb_47s_wb |>
  write_csv(file.path(root_fd, 'data', 'cn', 'ukb_47s_wb.csv'),
            na = '')
```


## GWAS Covariates

### White British

```{r}
ukb_gwas_covars_wb <- ukb_demo |>
  inner_join(ukb_bm, by = 'eid') |>
  inner_join(ukb_geno_array_wb, by = 'eid') |>
  filter(Array == 1) |>
  inner_join(ukb_pcs, by = 'eid') |>
  inner_join(ukb_blood_pc, by = 'eid') |>
  inner_join(ukb_sr, by = 'eid') |>
  inner_join(ukb_atl, by = 'eid') |>
  inner_join(ukb_47s_wb, by = 'eid') |>
  inner_join(ukb_ac |> select(-coding), by = 'eid') |>
  mutate(`Age Sq` = Age * Age) |>
  select(eid, Sex,
         starts_with('Age'), 
         contains('Centre'),
         paste0('PC', 1:10),
         starts_with('Blood'),
         starts_with('Adj'),
         CN, Height) 
```

```{r}
ukb_gwas_covars_wb |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_gwas_covars.csv'),
  na = '')
```

```{r}
ukb_gwas_covars_wb_blmm <- ukb_gwas_covars_wb |> 
  mutate(FID=eid, IID=eid) |>
  select(-eid) |>
  relocate(FID, .before=Sex) |>
  relocate(IID, .before=Sex) |>
  mutate(Sex = as.numeric(as.factor(Sex)),
         `As Centre` = as.numeric(as.factor(`As Centre`)),
         `Seq Centre` = as.numeric(as.factor(`Seq Centre`)))

colnames(ukb_gwas_covars_wb_blmm) <- gsub(' ', '_',
       c('FID', 'IID',
         tolower(
           colnames(
             ukb_gwas_covars_wb_blmm |> select(-ends_with('ID'))))))
```

```{r}
ukb_gwas_covars_wb_blmm |>
  write_delim(file.path(root_fd, 'sc_rap', 'rap_gwas_imp', 'covar.txt'),
              delim=' ')
```



```{r}
ukb_gwas_covars_o <- ukb_demo |>
  inner_join(ukb_bm, by = 'eid') |>
  inner_join(ukb_geno_array_wb, by = 'eid') |>
  filter(Array == 1) |>
  inner_join(ukb_pcs, by = 'eid') |>
  inner_join(ukb_blood_pc, by = 'eid') |>
  inner_join(ukb_sr, by = 'eid') |>
  inner_join(ukb_atl, by = 'eid') |>
  inner_join(ukb_47s_wb, by = 'eid') |>
  inner_join(ukb_ac |> select(-coding), by = 'eid') |>
  mutate(`Age Sq` = Age * Age) |>
  select(eid, Sex,
         starts_with('Age'), 
         contains('Centre'),
         paste0('PC', 1:10),
         starts_with('Blood'),
         starts_with('Adj'),
         CN, Height) 
```

### deCODE

```{r}
ukb_gwas_covars_decode_blmm <- ukb_gwas_covars_wb |> 
  filter(`Seq Centre` == 'deCODE') |>
  mutate(FID=eid, IID=eid) |>
  select(-eid, -`Seq Centre`) |>
  relocate(FID, .before=Sex) |>
  relocate(IID, .before=Sex) |>
  mutate(Sex = as.numeric(as.factor(Sex)),
         `As Centre` = as.numeric(as.factor(`As Centre`)))

colnames(ukb_gwas_covars_decode_blmm) <- gsub(' ', '_',
       c('FID', 'IID',
         tolower(
           colnames(
             ukb_gwas_covars_decode_blmm |> select(-ends_with('ID'))))))
```

```{r}
ukb_gwas_covars_decode_blmm |>
  write_delim(
    file.path(root_fd, 'sc_rap', 'rap_gwas_imp', 'covar_decode.txt'),
    delim=' ')
```

### Sanger

```{r}
ukb_gwas_covars_sanger_blmm <- ukb_gwas_covars_wb |> 
  filter(`Seq Centre` == 'Sanger') |>
  mutate(FID=eid, IID=eid) |>
  select(-eid, -`Seq Centre`) |>
  relocate(FID, .before=Sex) |>
  relocate(IID, .before=Sex) |>
  mutate(Sex = as.numeric(as.factor(Sex)),
         `As Centre` = as.numeric(as.factor(`As Centre`)))

colnames(ukb_gwas_covars_sanger_blmm) <- gsub(' ', '_',
       c('FID', 'IID',
         tolower(
           colnames(
             ukb_gwas_covars_sanger_blmm |> select(-ends_with('ID'))))))
```

```{r}
ukb_gwas_covars_sanger_blmm |>
  write_delim(
    file.path(root_fd, 'sc_rap', 'rap_gwas_imp', 'covar_sanger.txt'),
    delim=' ')
```

## Smokers

```{r}
ukb_smokers <- ukb_pheno_initial |> 
  select(eid, smoking=starts_with('20116-0.0')) |>
  mutate(Smoker = forcats::fct_rev(if_else(
    is.na(smoking) | smoking < 0 | smoking > 1,
    'Currently or Unknown',
    'Not Currently'
  ))) |>
  select(-smoking)
```

```{r}
ukb_smokers |>
  write_csv(
    file.path(root_fd, 'data', 'pheno', 'ukb_smokers.csv'),
    na = '')
```

## Congenital Diseases

```{r warning = F}
ukb_cong <- read_csv(
  file.path(root_fd, 'data', 'metadata', 'ukb671089.csv'),
                     col_types = cols()) |>
  select(eid, ends_with('-0.0')) |>
  select(eid, 
         starts_with('1324'),
         starts_with('1325'),
         starts_with('1326')) |>
  filter(eid %in% ukb_pheno_last$eid)
```

```{r}
bind_cols(
  ukb_cong |> select(eid), 
  ukb_cong |> select_if(lubridate::is.Date)) |>
  group_by(eid) |>
  pivot_longer(-eid, names_to = 'disease', values_to = 'date') |>
  summarise(num_diseases = sum(!is.na(date))) |>
  write_csv(
    paste0(root_fd, 'data', 'pheno', 'ukb_cong.csv'),
    na = '')
```

## Cancer

```{r}
ukb_cancer_codes <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'var_20001_enc.csv'),
  col_types = cols()) |>
  select(-selectable, -blood)

ukb_cancer <- read_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_diseases.csv'),
                         col_types = cols()) |>
  filter(eid %in% ukb_pheno_last$eid) |>
  filter(var_id == 20001, instance == 0) |>
  group_by(eid) |> summarise(
    Cancer = if_else(any(disease != 0), 'Yes', 'No'), 
    .groups = 'drop')
```

```{r}
ukb_cancer |> write_csv(
  file.path(root_fd, 'data', 'pheno', 'ukb_infl.csv'),
  na = '')
```
